---
title: "Predict props to cells per flag, year, gear, length using the regional models"
output: html_document
date: "2025-07-29"
---

# Summary

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(terra)
library(foreach)
library(doParallel)
library(progress)
library(glue)
library(qs)
library(here)
library(janitor)
library(arrow)
library(countrycode)
library(sf)
library(strex)
library(ranger)

source(here("R/dir.R"))
```

## Functions to prepare enviro and effort data  

```{r}
elnino <- read.csv(here("data/model_features/enso_index.csv"))

pdo <- read.csv(here("data/model_features/pdo_index.csv")) 

world_bank <- read.csv(here("data/model_features/world_bank_regions.csv")) 

gfi_df <- read.csv(here("data/model_features/global_fishing_index_governance.csv"))

# Load prepared data
model_data <- readRDS(here("data/model_features/prepared_data_0.5.rds")) %>%
  mutate(year = as.numeric(year)) %>%
  mutate(presence = 1) 

## read in environmental variables

  global_grid <- read.csv(here("data/model_features/global_grid.csv"))
  
  chl_data <- qs::qread(here("data/model_features/chl_data.qs"))
  
  sst_data <- read.csv(here("data/model_features/sst_data.csv"))

  wind_data <- qs::qread(here("data/model_features/wind_data.qs"))

  ocean_data <- sst_data %>%
    left_join(chl_data) %>%
    left_join(wind_data) %>%
    filter(year %in% c(2015:2017))
  
ocean_data_historical <- qs::qread(here("data/int/prediction_historical_data/chl_yearly_1950_2014.qs")) %>%
  left_join(qs::qread(here("data/int/prediction_historical_data/sst_yearly_1950_2014.qs")), by = c("pixel_id", "year")) %>% 
  left_join(qs::qread(here("data/int/prediction_historical_data/wind_yearly_1950_2014.qs")) %>% mutate(year = as.numeric(year)), by = c("pixel_id", "year"))

  ocean_data_all <- ocean_data %>%
    rbind(ocean_data_historical)

  spatial_data <- global_grid %>%
    left_join(read.csv(here("data/model_features/gfw_static_spatial_measures.csv")) %>% dplyr::select(-lat, -lon), by = "pixel_id") %>%
   left_join(read.csv(here("data/model_features/mesopelagiczones/mesopelagiczones_fixed.csv")), by = "pixel_id") %>%
   left_join(read.csv(here("data/model_features/eez/eez.csv")), by = "pixel_id") %>%
   left_join(read.csv(here("data/model_features/fao/fao_fixed.csv")), by = "pixel_id") %>%
   left_join(read.csv(here("data/model_features/ocean/ocean_fixed.csv")), by = "pixel_id") %>%
    left_join(read.csv(here("data/model_features/seamounts.csv")), by = "pixel_id") %>% 
   crossing(., year = c(1950:2017)) 
  
  
  env_data <- spatial_data %>%
    left_join(ocean_data_all) %>% # cool.
    left_join(elnino) %>%
    left_join(pdo) %>%
    left_join(world_bank, by = c("eez_id")) %>%
    mutate(eez_region_world_bank_7 = ifelse(eez_sovereign %in% c("High seas", "Land"), "High seas", eez_region_world_bank_7)) %>% 
    left_join(gfi_df, by = c("eez_sovereign" = "flag_fin")) %>% # add in global fishing index data here
    mutate(gov_score = ifelse(eez_id >= 99999 & is.na(gov_score), "high_seas", gov_score)) %>%
        mutate(gov_score = ifelse(eez_id < 99999 & is.na(gov_score), "no_data", gov_score)) %>%
    dplyr::select(-eez_sovereign, -nearest_seamount_id, -geometry_wkt) %>% 
    distinct()
  

mollweide_projection <- "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +units=m +no_defs"

data_grid_area <- global_grid %>%
    st_as_sf(wkt = "geometry_wkt",
           crs = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")%>% 
  st_wrap_dateline(options = c("WRAPDATELINE=YES", "DATELINEOFFSET=180"), quiet = TRUE) %>% 
  st_transform(mollweide_projection) %>%
  mutate(pixel_area_m2 = st_area(geometry_wkt)%>%
           units::drop_units()) 

```

```{r}

rousseau_regions <- read.csv("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/SAUPtoCountry.csv") %>%
  dplyr::select(-X) %>%
  rename(region = Region) 

fix_landlocked <- rousseau_regions %>%
  filter(region == "Landlocked") %>%
  mutate(region = case_when(
    Country == "AFG" ~ "MidEast",
    Country == "AND" ~ "Europe",
    Country == "AZE" ~ "MidEast",
    Country == "AUT" ~ "Europe",
    Country == "ARM" ~ "MidEast",
    Country == "BTN" ~ "SouthEastAsia",
    Country == "BOL" ~ "LatinAmerica",
    Country == "BWA" ~ "SubAfrica",
    Country == "BDI" ~ "SubAfrica",
    Country == "BLR" ~ "Europe",
    Country == "CAF" ~ "SubAfrica",
    Country == "TCD" ~ "NWAfrica",
    Country == "CSK" ~ "Europe",   # Historical, treat as Europe if included
    Country == "CZE" ~ "Europe",
    Country == "ETH" ~ "SubAfrica",
    Country == "HUN" ~ "Europe",
    Country == "KAZ" ~ "MidEast",
    Country == "KGZ" ~ "MidEast",
    Country == "LAO" ~ "SouthEastAsia",
    Country == "LSO" ~ "SubAfrica",
    Country == "LIE" ~ "Europe",
    Country == "LUX" ~ "Europe",
    Country == "MWI" ~ "SubAfrica",
    Country == "MLI" ~ "NWAfrica",
    Country == "MNG" ~ "NorthEastAsia",
    Country == "MDA" ~ "Europe",
    Country == "NPL" ~ "SouthEastAsia",
    Country == "NER" ~ "NWAfrica",
    Country == "PRY" ~ "LatinAmerica",
    Country == "RWA" ~ "SubAfrica",
    Country == "SMR" ~ "Europe",
    Country == "SVK" ~ "Europe",
    Country == "SWZ" ~ "SubAfrica",
    Country == "CHE" ~ "Europe",
    Country == "TJK" ~ "MidEast",
    Country == "TKM" ~ "MidEast",
    Country == "UGA" ~ "SubAfrica",
    Country == "MKD" ~ "Europe",
    Country == "BFA" ~ "NWAfrica",
    Country == "UZB" ~ "MidEast",
    Country == "SRB" ~ "Europe",
    Country == "ZMB" ~ "SubAfrica",
    TRUE ~ region
  ))

rousseau_fix <- rousseau_regions %>%
  filter(region != "Landlocked") %>%
  rbind(fix_landlocked) %>%
  dplyr::select(-SAUP) %>%
  filter(region  != "Unknown") %>%
    add_row(Country = "ATF", region = "IslandNations") %>%
    add_row(Country = "VAT", region = "Europe") %>%
    add_row(Country = "BES", region = "Caribbean") %>%
    add_row(Country = "CCK", region = "IslandNations")

hist_fish_data <- qs::qread(here("data/model_features/rousseau_total_art_vessels.qs")) %>%
  dplyr::select(-region) %>%
    filter(nv > 0) %>%
  mutate(flag_country_name = countrycode(flag_fin, origin = "iso3c", destination = "country.name")) %>%
  mutate(flag_country_name = 
           case_when(
             flag_fin == "RAA" ~ "Azores", 
             flag_fin == "RAM"  ~ "Madeira",
             TRUE ~ flag_country_name
           )) %>%
  left_join(rousseau_fix, by = c("flag_fin" = "Country"))

  
flags <- unique(hist_fish_data$flag_fin) # get the flags we need to run models for # 147 of them

env_grid <- env_data %>% 
  dplyr::select(lon, lat) %>% distinct()

```


Read in EEZ id and FAO id lookup tables so we can save with the actual iso3c and FAO id numbers

```{r}

eez_lookup <- read.csv(here("data/model_features/eez/eez_lookup.csv")) %>%
  clean_names() %>%
  mutate(eez_country_name = countrycode(eez_sovereign, origin = "iso3c", destination = "country.name")) %>%
  mutate(eez_country_name = ifelse(is.na(eez_country_name), "High seas", eez_country_name)) %>%
  mutate()

fao_lookup <- read.csv(here("data/model_features/fao/fao_major_ids.csv")) %>%
  clean_names() %>%
  dplyr::select(fao_id, "fao_major_fishing_area" = "name_en")

global_grid <- read.csv(here("data/model_features/global_grid.csv")) %>%
  dplyr::select(lon, lat, pixel_id)

```
Load access data 

```{r}

access_data_loop <- qs::qread(here("data/int/prediction_historical_data/art_fishing_access.qs")) %>%
  distinct()

env_grid <- env_data %>% 
  dplyr::select(lon, lat) %>% distinct()
```


After the models are fit, we make predictions on ALL DATA. So stage 1 is in sample predictions (since we fit the stage one model on all data), and stage two will contain out of sample predictions (since we fit the model on only data with fishing effort in it!)

 - This gapfilling fills ~10% of global production 
 
```{r}
years <- c(1950:2017)

# what region models do we have? 
model_data_loop <- model_data %>%
  left_join(rousseau_fix, by = c("flag_fin" = "Country")) %>%
  filter(!is.na(region))

flags_model <- unique(model_data_loop$flag_fin)

regions <- c(unique(model_data_loop$region))

flags <- hist_fish_data %>%
  filter(region %in% regions) %>%
  pull(flag_fin) %>%
  unique() # 121 can be filled in through this 
 

cl <- makeCluster(30)  
registerDoParallel(cl)
foreach(flag = flags, .packages = c("dplyr", "tidyverse", "randomForest", "qs", "glue", "ranger")) %dopar% {
# foreach(yr = years, .packages = c("dplyr", "tidyverse", "randomForest", "qs", "glue", "ranger")) %dopar% {
 for(yr in years) {
# flag = "AGO"
# yr = 2017
    
    # if(file.exists(file.path(glue("/home/ubuntu/data_storage/prep/random_forest/minderoo_artisanal/predictions_regional/yearly/model_preds_{flag}_{yr}.qs")))){
    #   cat("skipping... exists")
    #   next()
    # }
    
# get combination of categories we need to run through our models
model_data_flag <- hist_fish_data %>%
  dplyr::select(flag_fin, length_category, year, nv, region) %>%
  filter(flag_fin == flag, year == yr)
  
distinct_cats <- model_data_flag %>%
    distinct(flag_fin, length_category) 

full_grid <- tidyr::crossing(env_grid, distinct_cats) %>%
  crossing(., year = yr)
  
full_data <- full_grid %>%
  left_join(env_data, by = c("lon", "lat", "year")) %>%
  dplyr::select(-pixel_id) 

if(nrow(full_data) == 0){
    cat("skipping", flag, yr, "check to make sure this is right")
  next()
}

region_loop <- model_data_flag %>%
  pull(region) %>%
  unique()

# read in stage 2 model! 
# Get the list of matching files
matching_files <- list.files(
  path = file.path(rdsi_dir, "prep/random_forest/artisanal_minderoo/stage_2_models_regional"), 
  pattern = glue::glue("stage_2_rf_model_full_data_{region_loop}_.*\\.qs$"),
  full.names = TRUE
)

# Check if a matching file exists
if (length(matching_files) > 0) {
  stage2_model <- qs::qread(matching_files[1])  # Read the first matching file
} else {
  stop("No matching file found for flag: ", flag)
}

set.seed(123)
oos_preds <- full_data %>%
  mutate(pred_prop = predict(stage2_model, data = .)$predictions)

## read in stage 1 model! 
# Get the list of matching files
matching_files <- list.files(
  path = file.path(rdsi_dir, "prep/random_forest/artisanal_minderoo/stage_1_models_regional/"), 
  pattern = glue::glue("stage_1_rf_model_full_data_{region_loop}_.*\\.qs$"),
  full.names = TRUE
)

# Check if a matching file exists
if (length(matching_files) > 0) {
  stage1_model <- qs::qread(matching_files[1])  # Read the first matching file
} else {
  stop("No matching file found for flag: ", flag)
}

set.seed(123)
## make stage 1 predictions and apply fishing access filter 
stage1_preds <- full_data %>%
  mutate(pred_presence = predict(stage1_model, data = .)$predictions[, "1"]) %>%
  mutate(pred_presence = ifelse(is.na(pred_presence), 0, pred_presence)) %>%
  left_join(eez_lookup) %>%
  left_join(access_data_loop %>% dplyr::select(-eez_sovereign)) %>%
  mutate(access = ifelse(is.na(access), 0, access)) %>% 
  mutate(access = ifelse(eez_sovereign == flag, 1, access)) %>% # only allow fishing in their EEZ
  mutate(pred_presence_access = access*pred_presence)

# now rescale predictions so that if the preds are > 0 and in their EEZ, then make them 1. 
stage_1_preds_rescale <- stage1_preds %>%
  left_join(eez_lookup) %>%
  mutate(pred_presence_rescale = ifelse(pred_presence_access > 0 & flag_fin == eez_sovereign, 1, 0)) 


## join stage 1 and 2 preds together and multiply 
all_preds <- oos_preds %>%
   left_join(., stage_1_preds_rescale) %>% 
   dplyr::select(lon, lat, flag_fin, length_category, year, fao_id, eez_id, eez_sovereign, pred_prop, pred_presence_rescale) %>% 
  mutate(pred_prop_final = pred_prop*pred_presence_rescale) %>%
  filter(pred_prop_final > 0) %>%
  filter(!is.na(pred_prop_final)) %>%
 group_by(year, flag_fin, length_category) %>%
  mutate(prop_nv_cell_predict_rescaled = pred_prop_final / sum(pred_prop_final, na.rm = TRUE)) %>%
 ungroup() %>% # need to rescale the predictions to be between 0 and 1 here so that we can allocate effort
  left_join(hist_fish_data) %>%
  mutate(nv = prop_nv_cell_predict_rescaled*nv) %>%
    mutate(nom_active_fishing_days = nv*days_at_sea_mean*gt_kw_mean) %>% # this is days right? yes
  mutate(eff_active_fishing_days = nom_active_fishing_days * (1.0261 ^ (year - 1950))) %>% # now calculate effective effort, using the formula from Rousseau et al. 2019
  left_join(global_grid) %>%
  left_join(fao_lookup) %>%
  left_join(eez_lookup) %>%
    dplyr::select(pixel_id, lon, lat, year, flag_fin, length_category, eez_sovereign, fao_id, fao_major_fishing_area, nv) %>%
  filter(nv > 0) %>%
  mutate(sector = "Artisanal")

if(nrow(all_preds) == 0){
    cat("skipping", flag, yr, "no predictions made...")
  next()
}

# test_preds <- all_preds %>%
#   # filter(eez_sovereign == "USA") %>%
#   dplyr::group_by(lon, lat) %>%
#   summarise(nv = sum(nv, na.rm = TRUE)) %>%
#   ungroup() %>%
#   rast(., type = "xyz")
# 
# plot(test_preds)
# plot(log(test_preds+1)) # Cool!


qs::qsave(all_preds, file.path(rdsi_dir, glue("prep/random_forest/artisanal_minderoo/predictions_regional/yearly/model_preds_{flag}_{yr}.qs"))) # qs is smaller

  }
}

stopCluster(cl)


```

Combine all years for flag and save 

```{r}
data_grid_area <- data_grid_area %>%
st_drop_geometry()

cl <- makeCluster(30) 
registerDoParallel(cl)

foreach(flag = flags, .packages = c("qs", "dplyr", "tidyverse", "glue", "countrycode")) %dopar% {
  # flag = "UKR"
  
  all_files_flag <- list.files(file.path(rdsi_dir, "prep/random_forest/artisanal_minderoo/predictions_regional/yearly/"), pattern = flag, full.names = TRUE)
  
    # Skip iteration if no files are found
  if (length(all_files_flag) == 0) next
  
  all_data_flag <- lapply(all_files_flag, qread) %>%
    bind_rows() %>%
    left_join(eez_lookup) %>%
    mutate(flag_country_name = countrycode(flag_fin, origin = "iso3c", destination = "country.name")) %>%
        mutate(eez_country_name = countrycode(eez_sovereign, origin = "iso3c", destination = "country.name")) %>%
    mutate(eez_country_name = ifelse(eez_sovereign == "High seas", "High seas", eez_country_name)) %>%
    dplyr::select(pixel_id, lon, lat, year, flag_fin, flag_country_name, length_category, eez_sovereign, eez_country_name, fao_id, fao_major_fishing_area, nv, sector) %>%
        rename(flag_country_iso3c = flag_fin, eez_sovereign_iso3c = eez_sovereign, eez_sovereign_name = eez_country_name, fao_fishing_id = fao_id) %>%
    left_join(data_grid_area) %>%
    mutate(pixel_area_km2 = pixel_area_m2/1000000) %>%
    mutate(nv_km2 = nv/pixel_area_km2) # calculate hours per km2 here too
  
  qs::qsave(all_data_flag, file.path(rdsi_dir, glue("prep/random_forest/artisanal_minderoo/predictions_regional/model_preds_1950_2017_{flag}.qs")))
}

stopCluster(cl)



  all_files_flag <- str_before_first(str_after_last(list.files(file.path(rdsi_dir, "prep/random_forest/artisanal_minderoo/predictions_regional/"), pattern = ".qs", full.names = TRUE), "_"), "\\.qs")
  
  setdiff(flags, all_files_flag) # 0 - good

```

