---
title: "Aggregate vessel and overpass data from Skylight"
output: html_document
date: "2025-07-22"
---

## Summary

Here we aggregate Skylight data to 0.5 degree resolution and determine a proxy of overpasses from the images.csv files

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(terra)
library(dplyr)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(qs)
library(janitor)
library(data.table)
library(lwgeom)
library(RANN) 

source(here("R/dir.R"))

data_grid_raw <- read.csv(here("data/model_features/global_grid.csv"))

data_grid <- data_grid_raw %>%
  st_drop_geometry() %>%
  dplyr::select(pixel_id, lon, lat)

vessels_final <- qs::qread(here("data/vessels/all_batch_lengths_year.qs"))
```

1. bin Skylight detections (using centroids) to 0.5° and counts D by (lon,lat,year,month,length_class),

2. builds image polygons from image.csv (all batches),

3. counts looks = unique images intersecting each 0.5° bin per month/year, and

4. joins looks to detections to compute r = D/looks just like Sentinel-2.

5. fills any remaining D>0 & looks==0 from the 8-nearest neighbors (0.5° queen’s neighborhood)

6. Match to our data grid

```{r}
# =========================
# 0) INPUTS / ASSUMPTIONS
# =========================
# - 'vessels_final' has: centroid_lon, centroid_lat, fname, timestamp or date, and either:
#    vessel_length_m       (numeric; we’ll bin into length_class)
# - image CSVs live at:
#     file.path(rdsi_raw_dir, "minderoo_skylight/unzipped_1/image.csv")
#     file.path(rdsi_raw_dir, "minderoo_skylight/unzipped_2/image.csv")
#     file.path(rdsi_raw_dir, "minderoo_skylight/unzipped_3/image.csv")

# =========================
# 1) Detections -> 0.5° bins
# =========================
vf <- as.data.table(vessels_final)

# Month/year (canonical YYYY-MM-01 in UTC)
to_month <- function(x) format(as.POSIXct(x, tz = "UTC"), "%Y-%m-01")

# choose timestamp column
if ("timestamp" %in% names(vf)) {
  vf[, ts := as.POSIXct(timestamp, tz = "UTC")]
} else if ("date" %in% names(vf)) {
  vf[, ts := as.POSIXct(date, tz = "UTC")]
} else {
  stop("vessels_final must contain 'timestamp' or 'date'.")
}

vf[, `:=`(month = to_month(ts), year = format(ts, "%Y"))]


  # derive standard buckets
  breaks <- c(-Inf, 6, 12, 24, 50, Inf)
  labels <- c("Less than 6m","6-12m","12-24m","24-50m","Over 50m")
  vf[, length_class := cut(vessel_length_m, breaks = breaks, labels = labels, right = FALSE)]


# 0.5° binning (lower-left edges; matches earlier binning)
vf[, lon := -180 + floor((centroid_lon + 180) / 0.5) * 0.5 ]
vf[, lat :=  -90 + floor((centroid_lat  +  90) / 0.5) * 0.5 ]

# vf <- vf %>%
#   filter(is_in_water == "True")

# Aggregate detections per bin × month × length class
agg_05_sky <- vf[
  !is.na(lon) & !is.na(lat) & !is.na(length_class),
  .(D = .N),
  by = .(lon, lat, month, year, length_class)
]

# ==========================================
# 2) Image polygons -> 0.5° looks per month
# ==========================================
# Read image files (keep required columns)
img_paths <- file.path(rdsi_raw_dir, paste0("minderoo_skylight/unzipped_", 1:3), "image.csv")
imgs_list <- lapply(img_paths, function(p) {
  if (!file.exists(p)) return(NULL)
  as.data.table(read.csv(p))
})
imgs_list <- Filter(Negate(is.null), imgs_list)
keep_cols <- c("fname","wgs84_x1","wgs84_y1","wgs84_x2","wgs84_y2",
               "wgs84_x3","wgs84_y3","wgs84_x4","wgs84_y4","timestamp")
images_all <- rbindlist(lapply(imgs_list, function(dt) dt[, intersect(names(dt), keep_cols), with=FALSE]),
                        use.names = TRUE, fill = TRUE)

images_all <- images_all[timestamp != "unknown"]
images_all[, ts := as.POSIXct(timestamp, tz = "UTC")]
images_all[, `:=`(month = to_month(ts), year = format(ts, "%Y"))]

# Row-wise polygons
poly_list <- lapply(seq_len(nrow(images_all)), function(i) {
  r <- images_all[i]
  m <- matrix(c(r$wgs84_x1, r$wgs84_y1,
                r$wgs84_x2, r$wgs84_y2,
                r$wgs84_x3, r$wgs84_y3,
                r$wgs84_x4, r$wgs84_y4,
                r$wgs84_x1, r$wgs84_y1), ncol = 2, byrow = TRUE)
  st_polygon(list(m))
})
img_sf <- st_sf(
  fname = images_all$fname,
  month = images_all$month,
  year  = images_all$year,
  geometry = st_sfc(poly_list, crs = 4326)
) |>
  st_make_valid() |>
  st_buffer(1e-6)   # tiny buffer so edge-touching counts

# Build a grid that covers BOTH detections and images
pts_sf <- st_as_sf(unique(agg_05_sky[, .(lon, lat)]), coords = c("lon","lat"), crs = 4326, remove = FALSE)
bbox_union <- st_bbox(st_union(st_as_sfc(st_bbox(img_sf)), st_as_sfc(st_bbox(pts_sf))))
grid_sf <- st_make_grid(st_as_sfc(bbox_union), cellsize = c(0.5,0.5), what = "polygons", square = TRUE)
grid_sf <- st_sf(geometry = grid_sf)

# Bin labels via centroids using the SAME floor scheme
cent <- st_coordinates(st_centroid(grid_sf))
grid_sf$lon <- -180 + floor((cent[,1] + 180) / 0.5) * 0.5
grid_sf$lat <-  -90 + floor((cent[,2] +  90) / 0.5) * 0.5

# Intersections: unique images per cell × month
ix <- st_intersects(grid_sf, img_sf)
pairs <- data.table(
  cell_row = rep.int(seq_len(nrow(grid_sf)), lengths(ix)),
  img_row  = unlist(ix)
)
pairs[, `:=`(
  lon   = grid_sf$lon[cell_row],
  lat   = grid_sf$lat[cell_row],
  month = img_sf$month[img_row],
  year  = img_sf$year[img_row],
  fname = img_sf$fname[img_row]
)]
pairs <- unique(pairs, by = c("lon","lat","month","year","fname"))

looks_05 <- pairs[, .(looks = .N), by = .(lon, lat, month, year)]


length_levels <- c("Less than 6m","6-12m","12-24m","24-50m","Over 50m")

# Expand overpasses to all vessel length classes (effort is class-agnostic)
looks_05_expand <- looks_05[, .(length_class = length_levels), by = .(lon, lat, month, year, looks)]

# ====================================
# 3) Join looks -> detections; compute r
# ====================================
setkey(agg_05_sky, lon, lat, month, year, length_class)
setkey(looks_05,   lon, lat, month, year)

agg_05_sky_new <- looks_05[agg_05_sky]    # left-join looks onto per-length rows
agg_05_sky_new[is.na(looks), looks := 0L]

agg_05_sky_new <- agg_05_sky %>%
  full_join(looks_05_expand) %>%
  mutate(D = ifelse(is.na(D), 0, D),
         looks = ifelse(is.na(looks), 0, looks))

sum(agg_05_sky_new$D, na.rm = TRUE) # 193600
sum(agg_05_sky$D, na.rm = TRUE) # 193600

## ok now we need to gapfill any of the cells which have 0 looks but detections. we will gapfill with the nearest neighbor cells looks fro the same month/year

# Input: agg_05_sky_new with columns:
# lon, lat, month, year, length_class, D, looks  (0.5° bins; month as "YYYY-MM-01")
# helper
wrap180 <- function(x) ((x + 180) %% 360) - 180

# rows needing fill
need <- unique(agg_05_sky_new[D > 0 & looks == 0, .(lon, lat, month, year)])

  # length-agnostic lookup of looks
  L <- unique(agg_05_sky_new[, .(lon, lat, month, year, looks)])
  setkey(L, month, year, lon, lat)

  # 8-neighbor offsets (queen)
  offs <- CJ(dx = c(-0.5, 0, 0.5), dy = c(-0.5, 0, 0.5))[!(dx == 0 & dy == 0)]

  # cross-join need × offs via a dummy key 'k'
  need_nb <- need[, .(lon0 = lon, lat0 = lat, month, year)][, k := 1L]
  offs[, k := 1L]
  need_nb <- need_nb[offs, on = "k", allow.cartesian = TRUE][, k := NULL]

  # compute neighbor coords
  need_nb[, `:=`(
    lon = wrap180(lon0 + dx),
    lat =        (lat0 + dy)
  )][, c("dx","dy") := NULL]

  # join neighbor looks
  nb <- L[need_nb, on = .(month, year, lon, lat)]

  # max looks across neighbors (use median() if you prefer)
  looks_fill <- nb[, .(looks_nb = suppressWarnings(max(looks, na.rm = TRUE))),
                   by = .(lon0, lat0, month, year)]
  looks_fill[is.infinite(looks_nb), looks_nb := NA_integer_]

  # apply fills back to all length_class rows in that cell-month
  setkey(looks_fill, lon0, lat0, month, year)
  setkey(agg_05_sky_new, lon, lat, month, year)

  agg_05_sky_new <- looks_fill[agg_05_sky_new]
  agg_05_sky_new[
    D > 0 & looks == 0,
    looks := fifelse(!is.na(looks_nb) & looks_nb > 0, as.integer(looks_nb), 1L)
  ]
  agg_05_sky_new[, looks_nb := NULL]

agg_05_sky_new <- agg_05_sky_new %>%
  rename(lon = lon0, lat = lat0) %>%
  left_join(data_grid)


test <- agg_05_sky_new %>%
  filter(is.na(pixel_id)) %>%
  distinct(lon, lat) %>%
  mutate(sample = 1) %>%
  rast(., type = "xyz")

plot(test) # cool this is the sampling area

```

Ok, after joining with data_grid, we see that some pixel_ids are NA. This is because they are all very close to the shore. So we need to assign these cells to the closest pixel_id. 

```{r}

# put your data in a data.table
dt <- as.data.table(agg_05_sky_new)

# indices of rows with/without pixel_id
idx_missing <- which(is.na(dt$pixel_id))

  known <- dt[!is.na(pixel_id), .(lon, lat, pixel_id)]
  # build matrices of coordinates
  known_xy <- as.matrix(known[, .(lon, lat)])
  miss_xy  <- as.matrix(dt[idx_missing, .(lon, lat)])

  # 1-nearest neighbor for each missing row
  nn <- RANN::nn2(data = known_xy, query = miss_xy, k = 1)

  # fill pixel_id from nearest known neighbor
  dt[idx_missing, pixel_id := known$pixel_id[nn$nn.idx]]

  # optional sanity check
  message("NAs before: ", sum(is.na(agg_05_sky_new$pixel_id)),
          " | NAs after: ", sum(is.na(dt$pixel_id)))

# result in dt; if you want it back as data.frame:
agg_05_sky_new_filled <- as.data.frame(dt) %>%
  dplyr::select(-lon, -lat) %>%
  left_join(data_grid) %>%
  group_by(lon, lat, pixel_id, year, length_class) %>%
  summarise(D = sum(D, na.rm = TRUE), 
            looks = sum(looks, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(year = as.numeric(year))

qs::qsave(agg_05_sky_new_filled, here("data/detections_looks/skylight_detections_looks.qs"))


test <- qread(here("data/detections_looks/skylight_detections_looks.qs")) %>%
  group_by(year) %>%
  summarise(looks = sum(looks, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(data = "skylight")

files <- list.files(here("data/detections_looks/"), full.names = TRUE, pattern = "sentinel")

test_looks <- lapply(files, qread) %>%
  bind_rows() %>%
  group_by(year) %>%
  summarise(looks = sum(looks, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(data = "sentinel") %>%
  rbind(test)

# ok so looks are pretty consistent within each dataset. Skylight from 2009-2024 pretty consistent and sentinel from >2019 pretty consistent

```


