---
title: "Aggregate vessel data, match to EEZs, and calculate response variable for model"
output: html_document
date: "2025-07-22"
---

## Summary

Here we aggregate the Skylight data to 0.5 by 0.5 degree, matching our analysis grid


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(terra)
library(dplyr)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(qs)
library(janitor)
library(glue)
library(data.table)

source(here("R/dir.R"))

data_grid_raw <- read.csv(here("data/model_features/global_grid.csv"))

data_grid <- data_grid_raw %>%
  st_drop_geometry() %>%
  dplyr::select(pixel_id, lon, lat)
```

Read in vessel detection data with overpass information that we prepped 

```{r}
vessels_raw <- lapply(list.files(here("data/detections_looks/"), full.names = TRUE), qread) %>%
  bind_rows() %>%
  group_by(lat, lon, pixel_id, year, length_category = length_class) %>%
  summarise(D = sum(D, na.rm = TRUE),
            looks = sum(looks, na.rm = TRUE)) %>%
  ungroup()

test <- vessels_raw %>%
  filter(year == 2021) %>%
  distinct(lon, lat, looks) %>%
  rast(., type = "xyz")
plot(test) # looks good 

```

Read in raw Rousseau country level data

```{r}
# codes used in rousseau et al
country_codes <- read.csv("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/SAUPtoCountry.csv")

# country level effort data
rousseau_eff <- read.csv("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/TotalEffortby_FishingCountry_LengthBoat_Gear_Sector.csv") %>%
  filter(!is.na(Length_Category)) %>%
  dplyr::select(-X) %>%
  clean_names()
```


Figure out the flag country of each detection. We are assuming that the EEZ the vessel is fishing in is its flag country. 

         - “We make the reasonable assumption that the small-scale fishing sub-sector only operates within domestic waters (i.e., in their nation’s EEZ, <200 NM from shore). There are a few exceptions to this assumption, however, the total value of fish caught by this sub-sector in non-domestic waters is negligible [32].” https://www.sciencedirect.com/science/article/pii/S0308597X23001380 
         
         
```{r}
test <- vessels_raw %>%
  filter(is.na(pixel_id)) # make sure no NAs


# read in EEZ data 
eez_lookup <- read.csv(here("data/model_features/eez/eez_lookup_fix.csv"))
eez_df <- read.csv(here("data/model_features/eez/eez_fix.csv")) %>%
  left_join(eez_lookup)


vessel_agg_5_flag <- vessels_raw %>%
  left_join(eez_df) %>%
  mutate(flag_fin = eez_sovereign) # assume flag country = eez country

length(unique(vessel_agg_5_flag$eez_id)) # 171 flags

# fix hong kong cell to be only hong kong
test_hkg <- vessel_agg_5_flag %>%
  filter(pixel_id == 120609) %>%
  filter(eez_sovereign == "HKG")

vessel_agg_5_flag_fix <- vessel_agg_5_flag %>%
  filter(pixel_id != 120609) %>%
  rbind(., test_hkg) %>%
      dplyr::group_by(
      pixel_id, lon, lat,
      flag_fin, length_category,
      year
    ) %>%
  summarise(D = sum(D, na.rm = TRUE),
            looks = sum(looks, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(r = D/looks,
         presence = case_when(D > 0 ~ 1, 
                              D == 0 & looks > 0 ~ 0,
                              TRUE ~ NA)) 

test <- vessel_agg_5_flag_fix %>%
  filter(is.na(r))

```

Filter out vessel detections that are too big according to rousseau data

```{r}
# all_vessels <- qread(here("data/vessels/vessels_prop_05_deg_overpass_all.qs"))
sum(vessel_agg_5_flag_fix$D) # 40358849 this many detections before filtering

rousseau_df <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/TotalEffortby_FGroup_FishingCountry_LengthBoat_Sector.csv") %>%
  filter(Sector %in% c("APW", "UP"),
         NV > 0) %>%
  distinct(flag_fin = Country, length_category = Length_Category) # grab any definition of artisanal in each country since 1950

filtered_df <- rousseau_df %>%
  left_join(vessel_agg_5_flag_fix) # lose a lot of rows

sum(filtered_df$D) # 15692982; have only about 38% of the detections from before the filtering 
15692982/40358849

test <- filtered_df %>%
  filter(D > 0)

missing <- setdiff(rousseau_df$flag_fin, test$flag_fin) #[1] "COK" "GUM" "TUV" "STP" "HTI" "KNA" "BEL"; ok missing these seven countries because we excluded categories. For these countries, lets see if they have detections pre exclusion, and if they do, we will use those detections to inform their models

test2 <- vessel_agg_5_flag_fix %>%
  filter(flag_fin %in% c(missing, "BRB")) %>%
  filter(D > 0) %>%
  group_by(flag_fin, length_category) %>%
  summarise(D = sum(D)) %>%
  ungroup() # ok we will use the next closest length category for the missing countries to inform the models. 

unique(test2$flag_fin) # cool, there are detections now. Ok, now let's see if we can exclude any based on regional rules perhaps? 


fix_missing <- vessel_agg_5_flag_fix %>%
    filter(flag_fin %in% c(missing, "BRB"), 
           D > 0, 
           length_category %in% c("12-24m"))

filtered_df_fix <- filtered_df %>%
  rbind(fix_missing) %>%
        dplyr::group_by(
      pixel_id, lon, lat,
      flag_fin, length_category,
      year
    ) %>%
  summarise(D = sum(D, na.rm = TRUE),
            looks = sum(looks, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(r = D/looks,
         presence = case_when(D > 0 ~ 1, 
                              D == 0 & looks > 0 ~ 0,
                              TRUE ~ NA)) 

qs::qsave(filtered_df_fix, here("data/vessels/vessels_prop_05_deg_overpass_all.qs"))


unique(filtered_df$flag_fin)
```

