---
title: "Data wrangling - Inshore fishing area"
author: "Gage Clawson (IMAS)"
date: '`r format(Sys.time(), "%m/%d/%Y")`'
output: 
  pdf_document: 
    number_sections: yes
    toc: true
    toc_depth: 4
editor_options: 
  chunk_output_type: console
---

# Summary

Prep a inshore fishing area layer to apply to our predictions so that we can cut out cells that are unsuitable for fishing due to uninhabited coastlines. We consider cells which are <=50km from shore AND <=200m depth AND adjacent to a population (i.e., <=50km from a populated area)


from SAUP:
 > Thus, while industrial fisheries can have access to all waters within and outside EEZs, we define all small-scale fisheries as being restricted to more inshore, coastal waters. Therefore, we make use of the Inshore Fishing Area (IFA) concept as defined by Chuenpagdee et al. (2006), who defined IFA as: the area that extends from shore to either 50 km offshore or to the 200 m depth contour, whichever comes first.  We amend this definition slightly: only territories that are inhabited and have local fisheries are defined as having an IFA. We define inhabited areas as cells within 50km of a populated area, where the populated area has a population density > 1 person per km2 or the local population within a one-pixel radius is >= 200. Thus, uninhabited territories, such as many sub-Antarctic islands, which are not inhabited (or only host government employees as part of research or meteorological stations) do not have IFAs.



Population layer reference: 
Jan Volkholz, Stefan Lange, Inga Sauer, Christian Otto (2024): ISIMIP3b population input data (v1.2). ISIMIP Repository.  https://doi.org/10.48364/ISIMIP.889136.2


Download URL: https://www.isimip.org/gettingstarted/input-data-bias-adjustment/details/124/ 

1. Coastal land strip (inland only):
Take land polygons and keep just the band within, say, 50 km inland of the shoreline. This avoids counting towns far inland.

2. Populated coastal land (binary):
Mark a coastal land cell “populated” if either:

population density ≥ pmin_dens (e.g., 1 person/km²), or

local population within a one pixel radius (~20-60km?) ≥ 200 total population 
This combo suppresses tiny outposts but keeps small coastal towns.

3. Distance-to-populated-coast (over the sea):
Compute a distance transform from sea cells to the nearest populated coastal land cell. Sea cells with distance ≤ 50 km are considered “adjacent to inhabited coast”.

4. IFA inhabited mask:
IFA_inhabited = IFA_basic ∩ AdjacentToInhabitedCoast, where
IFA_basic = (distance_from_shore_km ≤ 50) OR (depth_m ≤ 200) on your fishing grid.



```{r echo = FALSE}
# This chunk sets up default settings for all chunks below
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,fig.width = 7.5,fig.height = 5,dev = 'png',dpi=300)
```

```{r include=FALSE}
# Load all necessary packages
library(tidyverse)
library(glue)
library(terra)
library(here)
library(sf)
library(rnaturalearth)
library(qs)
library(janitor)
library(tidyterra)
library(countrycode)

options(scipen = 20)

source(here("R/dir.R"))

data_directory <- rdsi_raw_dir

```

# Defining our global grid

```{r}
pixel_size <- 0.5

mollweide_projection <- "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +units=m +no_defs"

# Read in data_grid, generated in data_wrangling.Rmd
# We will do spatial joining in Mollweide, for proper calculations
# Note that data grid still retains lat and lon columns, for joining with non-spatial tibbles
# For Mollweide, always wrap around dateline, then transform, then calculate areas at end

data_grid <- data.table::fread(here::here("data/model_features/global_grid.csv")) %>%
  st_drop_geometry %>%
  dplyr::select(lon, lat, pixel_id)

data_rast <- rast(data_grid, type = "xyz", mollweide_projection)

```

Read in population data for 1950 to 2020. Luckily it is alread half degree. 
```{r}
# siconc monthly at 15' (units: 0–1; if %, divide by 100)
hist_nc  <- (file.path(data_directory, "ISIMIP/population/population_histsoc_30arcmin_annual_1901_2014.nc"))
future_nc <- (file.path(data_directory, "ISIMIP/population/population_2015soc_30arcmin_annual_2015_2100.nc"))

names(sds(hist_nc))
names(sds(future_nc))

# Read ONLY the "total-population" variable from each file
pop_hist   <- rast(hist_nc,   subds = "total-population")[[50:114]] # 1950 - 2010
pop_future <- rast(future_nc, subds = "total-population")[[1:10]] # 2015-2024

pop <- c(pop_hist, pop_future) %>%
  resample(data_rast, method = "bilinear") # should probably use this crs throughout


plot(pop[[75]]) # cool. 

years <- 1950:2024

names(pop) <- years

```

```{r}
## 2) Land / sea and projection for metric distances
# Use Mollweide for global distances (meters)
crs_proj <- "ESRI:54030"

land_sf <- rnaturalearth::ne_countries(scale = "small", returnclass = "sf")
land_v  <- vect(land_sf)

# Project population stack & land to metric CRS
pop_proj  <- project(pop, crs_proj, method = "near")  # counts -> use "near"
land_proj <- project(land_v, crs_proj)


# Rasterize land onto pop grid (single layer template)
tmpl <- pop_proj[[1]]
land_r <- rasterize(land_proj, tmpl, field = 1, background = NA)

# Sea mask (1 at sea, NA on land)
sea_r <- land_r
sea_r[] <- 1
sea_r[!is.na(land_r)] <- NA

```

```{r}
## 3) Coastal land band (≤ 50 km inland from shoreline)
# distance() to sea gives meters for every cell; mask to land ≤50 km
dist_to_sea_m <- distance(sea_r)

coastal_land_band <- dist_to_sea_m %>%
  filter(layer <= 50000) %>%
  mutate(layer = ifelse(layer > 0, 1, 0))


plot(coastal_land_band)

coastal_land_band <- mask(land_r, dist_to_sea_m <= 50000)

plot(dist_to_sea_m <= 50000)
plot(coastal_land_band)

```

```{r}
## 4) Population criteria on coastal land (per year)
# (a) density ≥ 1 person/km² OR (b) local population within 10 km ≥ 200
# Prepare area and focal window
area_km2 <- cellSize(tmpl, unit = "km")      # km² per cell (same for all layers at this res/CRS)
area_km2 <- rast(rep(list(area_km2), nlyr(pop_proj)))  # recycle to match layers
names(area_km2) <- names(pop_proj)

# create a zero raster with the same geometry so we can use cover
r0 <- pop_proj[[1]]
r0[] <- 0

# fill NAs in pop_proj with zeros from r0 
pop_proj <- cover(pop_proj, r0)

w <- matrix(1, nrow = 3, ncol = 3)
local_sum <- terra::focal(pop_proj, w = w, fun = "sum", na.rm = TRUE)

# Density (people per km²)
dens <- pop_proj / area_km2

# Inhabited if density≥1 OR local_sum≥200; then keep only in coastal land band
inhabited <- (dens >= 1) | (local_sum >= 200) # more than 200 people within a pixel (20-60km2) or more than 1 person per km2
inhabited_coast <- mask(inhabited, coastal_land_band)

```

FIX

```{r}
# inhabited land: 1 where inhabited, NA elsewhere
inhabited_land <- terra::classify(inhabited_coast, rbind(c(0, NA), c(1, 1)))

# land raster: 1 on land, NA on sea
land_r <- ifel(is.na(sea_r), 1, NA)

# identify coastal land cells: land cells with at least one neighboring sea cell
sea_neigh <- terra::focal(
  ifel(!is.na(sea_r), 1, 0),
  w = matrix(1, 3, 3),
  fun = sum,
  na.policy = "omit",
  fillvalue = 0
)

coast_land <- land_r
coast_land[sea_neigh == 0] <- NA   # keep only land touching sea

# for each year/layer, keep coastal land cells within 50 km of inhabited land
source_coast_list <- lapply(seq_len(nlyr(inhabited_land)), function(i) {
  inh <- inhabited_land[[i]]
  
  # distance to inhabited land, evaluated everywhere
  d_inh <- terra::distance(inh)
  
  # evaluate only on coastal land
  d_coast <- terra::mask(d_inh, coast_land)
  
  src <- d_coast <= 50000
  src[src == 0] <- NA
  names(src) <- paste0("y", years[i])
  src
})

source_coast <- terra::rast(source_coast_list)

writeRaster(source_coast, here("data/int/inhabited_coast_years.tif"), overwrite = TRUE)

```

```{r}
source_dist_list <- lapply(seq_len(nlyr(source_coast)), function(i) {
  src <- source_coast[[i]]           # coastal source cells for year i
  d   <- terra::distance(src)        # distance to source coast
  d_sea <- terra::mask(d, sea_r)     # keep distances over sea only
  names(d_sea) <- paste0("y", years[i])
  d_sea
})

source_dist <- terra::rast(source_dist_list)
writeRaster(source_dist, here("data/int/dist_to_inhabited_coast.tif"), overwrite = TRUE)

depth_df <- gfw_vars %>%
  dplyr::select(lon, lat, elevation_m)

depth_rast <- rast(depth_df, type = "xyz", crs = "EPSG:4326")

# project to same CRS/grid framework as sea_r
depth_proj <- project(depth_rast, sea_r, method = "near")

# if needed, force exact grid match
depth_align <- resample(depth_proj, sea_r, method = "near")

depth_sea <- mask(depth_align, sea_r)

dist_df <- gfw_vars %>%
  dplyr::select(lon, lat, distance_from_shore_m)

dist_rast <- rast(dist_df, type = "xyz", crs = "EPSG:4326")

# project to same CRS/grid framework as sea_r
dist_proj <- project(dist_rast, sea_r, method = "near")

# if needed, force exact grid match
dist_align <- resample(dist_proj, sea_r, method = "near")

dist_sea <- mask(dist_align, sea_r)


ifa_list <- lapply(seq_len(nlyr(source_dist)), function(i) {
  d <- source_dist[[i]]
  
  ifa <- (d <= 50000) & (depth_sea >= -200 | dist_sea <= 50)
  ifa <- terra::mask(ifa, sea_r)
  
  names(ifa) <- paste0("y", years[i])
  ifa
})

ifa_final <- terra::rast(ifa_list)
ifa_final_num <- as.numeric(ifa_final)

## ok now we need to put this on our analysis grid
ifa_fin   <- terra::project(ifa_final_num, crs(pop), method = "near")
ifa_fin_fin <- terra::resample(ifa_fin, pop, method = "average") # try average instead; bilinear too broad 
ifa_fin_fin <- terra::ifel(ifa_fin_fin >= 0.5, 1, 0) # make sure 

# ok now we need to extract to our data grid. From there we will gapfill any missing countries (because some countries are missing from the population data) with the basic IFA (e.g., doesn't matter if there is population on the coast or not)

ifa_fin_df <- as.data.frame(ifa_fin_fin, xy = TRUE)

ifa_fin_df_long <- ifa_fin_df %>% 
  rename(lon = x, lat = y) %>%
  pivot_longer(
    cols = starts_with("y"),
    names_to = "year",
    values_to = "ifa"
  ) %>%
  mutate(
    year = as.integer(sub("^y", "", year))
  )

test <- ifa_fin_df_long %>%
  filter(ifa > 0) %>%
  left_join(data_grid)


ifa_df_fin <- ifa_fin_df_long %>%
  left_join(data_grid)

```

Now we need to fix any NA pixel ids with nearest neighbor. For example, if pixel_id == NA and ifa ==1, we still want to retain that. 

```{r}
test <- ifa_df_fin %>%
  filter(ifa > 0,
         is.na(pixel_id), 
         year == 2017) %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") %>%
  plot() # so we want to keep all of these if possible. We will fill in with nearest neighbor pixel id. 


library(FNN)

# make a copy
ifa_df_fixed <- ifa_df_fin

#-----------------------------
# 1. Fill from exact lon/lat match first
#-----------------------------
coord_lookup <- ifa_df_fixed %>%
  filter(!is.na(pixel_id)) %>%
  distinct(lon, lat, pixel_id)

ifa_df_fixed <- ifa_df_fixed %>%
  left_join(coord_lookup, by = c("lon", "lat"), suffix = c("", "_lookup")) %>%
  mutate(
    pixel_id = if_else(is.na(pixel_id), pixel_id_lookup, pixel_id)
  ) %>%
  select(-pixel_id_lookup)

#-----------------------------
# 2. For remaining NA pixel_id, use nearest neighbor
#-----------------------------
na_idx <- which(is.na(ifa_df_fixed$pixel_id))
good_idx <- which(!is.na(ifa_df_fixed$pixel_id))

if (length(na_idx) > 0) {
  
  coords_good <- as.matrix(ifa_df_fixed[good_idx, c("lon", "lat")])
  coords_na   <- as.matrix(ifa_df_fixed[na_idx, c("lon", "lat")])
  
  nn <- get.knnx(
    data = coords_good,
    query = coords_na,
    k = 1
  )
  
  nearest_good_rows <- good_idx[nn$nn.index[, 1]]
  
  ifa_df_fixed$pixel_id[na_idx] <- ifa_df_fixed$pixel_id[nearest_good_rows]
}

# check
sum(is.na(ifa_df_fixed$pixel_id))


test <- ifa_df_fixed %>%
  filter(ifa > 0,
         year == 2017) %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") %>%
  plot() # so we want to keep all of these if possible. We will fill in with nearest neighbor pixel id. 

# perfect
```

Now fill in any missing countries. For the missing flag countries, we will just use <= 50km from coast OR 200m depth, doesn't have to be populated

```{r}

## now lets make sure that all of the EEZs are represented with an IFA. 
eez_lookup <- read.csv(here("data/model_features/eez/eez_lookup_fix.csv")) %>%
  clean_names() %>%
  mutate(eez_country_name = countrycode(eez_sovereign, origin = "iso3c", destination = "country.name")) %>%
  mutate(eez_country_name = case_when(
    eez_sovereign == "RAA" ~ "Azores", 
    eez_sovereign == "RAM" ~ "Madeira",
    TRUE ~ eez_country_name
  )) %>%
  mutate(eez_country_name = ifelse(is.na(eez_country_name), "High seas", eez_country_name))

eez <- read.csv(here("data/model_features/eez/eez_fix.csv"))

test <- ifa_df_fixed %>% 
  filter(year == 1990) %>%
  left_join(data_grid) %>%
  left_join(eez) %>%
  left_join(eez_lookup) %>%
  filter(eez_sovereign == "WLF") %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") %>%
  plot() # ok so all 0s for WLF. We don't want that.

ifa_eez <- ifa_fin_df_long %>% 
  filter(ifa == 1) %>%
  left_join(data_grid) %>%
  left_join(eez) %>%
  left_join(eez_lookup) 

missing_eez <- unique(setdiff(unique(eez_lookup$eez_sovereign), unique(ifa_eez$eez_sovereign)))

missing_eez
#  [1] "WSM" "TON" "KIR" "COM" "MUS" "SYC" "ATG" "BRB" "CPV" "DMA" "GRD" "KNA" "LCA" "VCT" "MLT" "MCO" "PSE" "STP" "BIH" "MHL" "PLW" "TUV" "FSM" "MDV" "NRU"
# [26] "COK" "PYF" "NIU" "MYT" "REU" "RAA" "FRO" "GLP" "RAM" "GUM" "WLF" "BMU"

sort(setdiff(unique(eez_lookup$eez_country_name), unique(ifa_eez$eez_country_name)))

## we are missing these from the IFA... ? Let's see if we can fix that. I assume it has something to do with the resolution of our population layer. These all have really small EEZs
#  [1] "Antigua & Barbuda"                "Azores"                           "Barbados"                         "Bermuda"                         
#  [5] "Bosnia & Herzegovina"             "Cape Verde"                       "Comoros"                          "Cook Islands"                    
#  [9] "Dominica"                         "Faroe Islands"                    "French Polynesia"                 "Grenada"                         
# [13] "Guadeloupe"                       "Guam"                             "Kiribati"                         "Madeira"                         
# [17] "Maldives"                         "Malta"                            "Marshall Islands"                 "Mauritius"                       
# [21] "Mayotte"                          "Micronesia (Federated States of)" "Monaco"                           "Nauru"                           
# [25] "Niue"                             "Palau"                            "Palestinian Territories"          "Réunion"                         
# [29] "Samoa"                            "São Tomé & Príncipe"              "Seychelles"                       "St. Kitts & Nevis"               
# [33] "St. Lucia"                        "St. Vincent & Grenadines"         "Tonga"                            "Tuvalu"                          
# [37] "Wallis & Futuna"    

ifa_fix_missing <- ifa_df_fin %>%
  left_join(data_grid) %>%
  left_join(eez) %>%
  left_join(eez_lookup) %>%
  filter(eez_sovereign %in% c(missing_eez))

test <- ifa_fix_missing %>%
  filter(year == 2017) %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") %>%
  plot() # ok, lets just exclude the population layer from these places and only filter for <200m or <50km 


  ifa_fix_missing_fin <- ifa_fix_missing %>%
      left_join(gfw_vars) %>%
  mutate(ifa = case_when(
    (elevation_m >= -200 | distance_from_shore_m <= 50) ~ 1, 
    TRUE ~ 0
  )) 
  
test <- ifa_fix_missing_fin %>%
  filter(year == 2017) %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") %>%
  plot() # looks good! 

test <- ifa_fix_missing_fin %>%
  filter(ifa == 1)
setdiff(missing_eez, unique(test$eez_sovereign)) # still missing MCO, PSE, and BIH. Are these actually in the Rousseau data? 

hist_fish_data <- qs::qread(here("data/model_features/rousseau_total_art_vessels.qs")) %>%
  mutate(flag_country_name = countrycode(flag_fin, origin = "iso3c", destination = "country.name")) %>%
  mutate(flag_country_name = 
           case_when(
             flag_fin == "RAA" ~ "Azores", 
             flag_fin == "RAM"  ~ "Madeira",
             TRUE ~ flag_country_name
           ))

  
flags <- as.data.frame(unique(hist_fish_data$flag_fin)) # get the flags we need to run models for # 151 of them

## ok cool, those missing are not actually in the Rousseau data, so we should be ok! 
ifa_fix_missing_fin <- ifa_fix_missing %>%
      left_join(gfw_vars) %>%
  mutate(ifa = case_when(
    (elevation_m >= -200 | distance_from_shore_m <= 50) ~ 1, 
    TRUE ~ 0
  )) %>%
  dplyr::select(pixel_id, lon, lat, year, ifa)
  
## now add these back into the ifa 
ifa_final_dataframe <- ifa_df_fin %>%
    left_join(eez) %>%
  left_join(eez_lookup) %>%
  filter(!(eez_sovereign %in% missing_eez)) %>%
  dplyr::select(pixel_id, lon, lat, year, ifa) %>%
  rbind(., ifa_fix_missing_fin)

qs::qsave(ifa_final_dataframe, here("data/model_features/ifa/ifa_fin.qs"))

test <- ifa_final_dataframe  %>%
  filter(year == 2017) %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") %>%
  plot() # looks good! 

```



ARCHIVE: 


```{r}
## 5) Sea-side adjacency: distance from sea cells to populated coastal land ≤ 50 km

# keep only truly inhabited coastal cells as sources (1 stays 1, 0 becomes NA)
inhabited_coast01 <- terra::classify(inhabited_coast, rbind(c(0, NA), c(1, 1)))

adjacent_list <- lapply(seq_len(nlyr(inhabited_coast01)), function(i) {
  src <- inhabited_coast01[[i]]          # NA except where inhabited==1
  d   <- terra::distance(src)            # meters to inhabited coastal land only
  d_sea <- terra::mask(d, sea_r)         # keep distances over sea
  adj <- d_sea <= 50000                  # within 50 km
  names(adj) <- paste0("y", years[i])
  adj
})
adjacent_stack_proj <- terra::rast(adjacent_list)
adjacent_stack_ll   <- terra::project(adjacent_stack_proj, crs(pop), method = "near")

```

```{r}
## 6) Extract to your ocean grid (lon/lat points)
pts <- vect(data_grid[, c("lon","lat")], geom = c("lon","lat"), crs = "EPSG:4326")

vals <- terra::extract(adjacent_stack_ll, pts)  # returns ID + one col per year-layer (logical)
vals$pixel_id <- data_grid$pixel_id

# Tidy to (pixel_id, year, adjacent_inhabited) long table
adj_long <- vals |>
  select(-ID) |>
  pivot_longer(-pixel_id, names_to = "layer", values_to = "adjacent_inhabited") |>
  mutate(year = as.integer(sub("^y", "", layer))) |>
  select(pixel_id, year, adjacent_inhabited) |>
  arrange(pixel_id, year) %>%
  mutate(adjacent_inhabited = ifelse(is.na(adjacent_inhabited) & !is.na(pixel_id), 0, adjacent_inhabited))


test <- adj_long %>% 
  filter(year == 2017) %>%
  left_join(data_grid) %>%
  dplyr::select(lon, lat, adjacent_inhabited) %>%
  rast(., type = "xyz") 
plot(test)


qs::qsave(adj_long, here("data/model_features/ifa/ifa_simple.qs"))

```

Now we need to read in depth and distance to shore and contain to <=50km OR <=200m depth cells 
 - Note: we are using OR because this is SAUP assumption and when I use AND it excludes some areas, like Wallis and Futuna. 

```{r}
ifa_simple <- qs::qread(here("data/model_features/ifa/ifa_simple.qs"))

test <- ifa_simple %>%
  left_join(data_grid) %>%
  filter(year == 2017) %>%
    dplyr::select(lon, lat, adjacent_inhabited) %>%
  rast(., type = "xyz") # ok so should look something like this 

plot(test)

gfw_vars <- read.csv(here("data/model_features/gfw_static_spatial_measures.csv")) %>%
  dplyr::select(pixel_id, lon, lat, distance_from_shore_m, elevation_m) # distance to shore is actually km, not m.

test <- gfw_vars %>%
  filter(distance_from_shore_m <= 50 | elevation_m >= -200) %>%
  mutate(ifa = 1) %>%
    dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") # ok so should look something like this 

test <- gfw_vars %>%
  filter(distance_from_shore_m <= 50) %>%
  mutate(ifa = 1) %>%
    dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") # ok so should look something like this 

plot(test)

ifa_fin <- ifa_simple %>% 
  left_join(gfw_vars) %>%
  mutate(ifa = case_when(
    (elevation_m >= -200 | distance_from_shore_m <= 50) & adjacent_inhabited == 1 ~ 1, 
    TRUE ~ 0
  )) %>%
  dplyr::select(pixel_id, lon, lat, year, ifa)

test <- ifa_fin %>%
  filter(year == 2017) %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz")


## now lets make sure that all of the EEZs are represented with an IFA. 
eez_lookup <- read.csv(here("data/model_features/eez/eez_lookup_fix.csv")) %>%
  clean_names() %>%
  mutate(eez_country_name = countrycode(eez_sovereign, origin = "iso3c", destination = "country.name")) %>%
  mutate(eez_country_name = case_when(
    eez_sovereign == "RAA" ~ "Azores", 
    eez_sovereign == "RAM" ~ "Madeira",
    TRUE ~ eez_country_name
  )) %>%
  mutate(eez_country_name = ifelse(is.na(eez_country_name), "High seas", eez_country_name))

eez <- read.csv(here("data/model_features/eez/eez_fix.csv"))

test <- ifa_fin %>% 
  filter(year == 1990 ) %>%
  left_join(data_grid) %>%
  left_join(eez) %>%
  left_join(eez_lookup) %>%
  filter(eez_sovereign == "WLF") %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") %>%
  plot()

ifa_eez <- ifa_fin %>% 
  left_join(data_grid) %>%
  left_join(eez) %>%
  left_join(eez_lookup) %>%
  filter(ifa == 1)

missing_eez <- unique(setdiff(unique(eez_lookup$eez_sovereign), unique(ifa_eez$eez_sovereign)))

#  [1] "KIR" "SYC" "KNA" "MLT" "TGO" "MCO" "PSE" "IRQ" "JOR" "BIH" "MHL" "PLW" "BEL" "TUV" "FSM" "MDV" "NRU" "COK" "PYF" "NIU" "RAA" "GUM"
# [23] "WLF" "BMU" "HKG"

sort(setdiff(unique(eez_lookup$eez_country_name), unique(ifa_eez$eez_country_name)))

## we are missing these from the IFA... ? Let's see if we can fix that. I assume it has something to do with the resolution of our population layer. These all have really small EEZs
#  [1] "Azores"                           "Belgium"                          "Bermuda"                          "Bosnia & Herzegovina"            
#  [5] "Cook Islands"                     "French Polynesia"                 "Guam"                             "Hong Kong SAR China"             
#  [9] "Iraq"                             "Jordan"                           "Kiribati"                         "Maldives"                        
# [13] "Malta"                            "Marshall Islands"                 "Micronesia (Federated States of)" "Monaco"                          
# [17] "Nauru"                            "Niue"                             "Palau"                            "Palestinian Territories"         
# [21] "Seychelles"                       "St. Kitts & Nevis"                "Togo"                             "Tuvalu"                          
# [25] "Wallis & Futuna" 


ifa_fix_missing <- ifa_fin %>%
    left_join(data_grid) %>%
  left_join(eez) %>%
  left_join(eez_lookup) %>%
  filter(eez_sovereign %in% c(missing_eez))

test <- ifa_fix_missing %>%
  filter(year == 2017) %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") %>%
  plot() # ok, lets just exclude the population layer from these places and only filter for <200m or <50km 


  ifa_fix_missing_fin <- ifa_fix_missing %>%
      left_join(gfw_vars) %>%
  mutate(ifa = case_when(
    (elevation_m >= -200 | distance_from_shore_m <= 50) ~ 1, 
    TRUE ~ 0
  )) 
  
test <- ifa_fix_missing_fin %>%
  filter(ifa == 1)

setdiff(missing_eez, unique(test$eez_sovereign)) # still missing MCO, PSE, and BIH. Are these actually in the Rousseau data? 

hist_fish_data <- qs::qread(here("data/model_features/rousseau_total_art_vessels.qs")) %>%
  mutate(flag_country_name = countrycode(flag_fin, origin = "iso3c", destination = "country.name")) %>%
  mutate(flag_country_name = 
           case_when(
             flag_fin == "RAA" ~ "Azores", 
             flag_fin == "RAM"  ~ "Madeira",
             TRUE ~ flag_country_name
           ))

  
flags <- as.data.frame(unique(hist_fish_data$flag_fin)) # get the flags we need to run models for # 151 of them

## ok cool, those missing are not actually in the Rousseau data, so we should be ok! 
ifa_fix_missing_fin <- ifa_fix_missing %>%
      left_join(gfw_vars) %>%
  mutate(ifa = case_when(
    (elevation_m >= -200 | distance_from_shore_m <= 50) ~ 1, 
    TRUE ~ 0
  )) %>%
  dplyr::select(pixel_id, lon, lat, year, ifa)
  
## now add these back into the ifa 
ifa_fin_fin <- ifa_fin %>%
  rbind(., ifa_fix_missing_fin)

qs::qsave(ifa_fin_fin, here("data/model_features/ifa/ifa_fin.qs"))


```