---
title: "Data wrangling - Inshore fishing area"
author: "Gage Clawson (IMAS)"
date: '`r format(Sys.time(), "%m/%d/%Y")`'
output: 
  pdf_document: 
    number_sections: yes
    toc: true
    toc_depth: 4
editor_options: 
  chunk_output_type: console
---

# Summary

Prep a inshore fishing area layer to apply to our predictions so that we can cut out cells that are unsuitable for fishing due to uninhabited coastlines. We consider cells which are <=50km from shore AND <=200m depth AND adjacent to a population (i.e., <=50km from a populated area)


from SAUP:
 > Thus, while industrial fisheries can have access to all waters within and outside EEZs, we define all small-scale fisheries as being restricted to more inshore, coastal waters. Therefore, we make use of the Inshore Fishing Area (IFA) concept as defined by Chuenpagdee et al. (2006), who defined IFA as: the area that extends from shore to either 50 km offshore or to the 200 m depth contour, whichever comes first.  We amend this definition slightly: only territories that are inhabited and have local fisheries are defined as having an IFA. Thus, uninhabited territories, such as many sub-Antarctic islands, which are not inhabited (or only host government employees as part of research or meteorological stations) do not have IFAs. The areas in km2 are measured similar to EEZ and shelf areas.



Population layer reference: 
Jan Volkholz, Stefan Lange, Inga Sauer, Christian Otto (2024): ISIMIP3b population input data (v1.2). ISIMIP Repository.  https://doi.org/10.48364/ISIMIP.889136.2


Download URL: https://www.isimip.org/gettingstarted/input-data-bias-adjustment/details/124/ 

1. Coastal land strip (inland only):
Take land polygons and keep just the band within, say, 50 km inland of the shoreline. This avoids counting towns far inland.

2. Populated coastal land (binary):
Mark a coastal land cell “populated” if either:

population density ≥ pmin_dens (e.g., 1 person/km²), or

local population within a one pixel radius (~20-60km?) ≥ 200 total population 
This combo suppresses tiny outposts but keeps small coastal towns.

3. Distance-to-populated-coast (over the sea):
Compute a distance transform from sea cells to the nearest populated coastal land cell. Sea cells with distance ≤ 50 km are considered “adjacent to inhabited coast”.

4. IFA inhabited mask:
IFA_inhabited = IFA_basic ∩ AdjacentToInhabitedCoast, where
IFA_basic = (distance_from_shore_km ≤ 50) OR (depth_m ≤ 200) on your fishing grid.



```{r echo = FALSE}
# This chunk sets up default settings for all chunks below
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,fig.width = 7.5,fig.height = 5,dev = 'png',dpi=300)
```

```{r include=FALSE}
# Load all necessary packages
library(tidyverse)
library(glue)
library(terra)
library(here)
library(sf)
library(rnaturalearth)
library(qs)
library(tidyterra)
library(janitor)
library(countrycode)

options(scipen = 20)

source(here("R/dir.R"))

data_directory <- rdsi_raw_dir

```

# Defining our global grid

```{r}
pixel_size <- 0.5

mollweide_projection <- "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +units=m +no_defs"


data_grid <- data.table::fread(here::here("data/model_features/global_grid.csv")) %>%
  st_drop_geometry %>%
  dplyr::select(lon, lat, pixel_id)

```

Read in population data for 1950 to 2020. Luckily it is already half degree. 
```{r}
# siconc monthly at 15' (units: 0–1; if %, divide by 100)
hist_nc  <- (file.path(data_directory, "ISIMIP/population/population_histsoc_30arcmin_annual_1901_2014.nc"))
future_nc <- (file.path(data_directory, "ISIMIP/population/population_2015soc_30arcmin_annual_2015_2100.nc"))

names(sds(hist_nc))
names(sds(future_nc))

# Read ONLY the "total-population" variable from each file
pop_hist   <- rast(hist_nc,   subds = "total-population")[[50:114]] # 1950 - 2010
pop_future <- rast(future_nc, subds = "total-population")[[1:10]] # 2015-2024

pop <- c(pop_hist, pop_future) %>%
  resample(data_rast, method = "bilinear")


plot(pop[[1]]) # cool. 

years <- 1950:2024

names(pop) <- years

data_rast <- rast(data_grid, type = "xyz", crs = pop[[1]])


```

Mask population layer to only have cells within 50km of the coast

```{r}
ocean_rast <- rast(here("data/model_features/ocean/ocean_rast.tif")) %>%
  mutate(name = ifelse(!is.na(name), 1, NA)) %>%
  disagg(., fact = 4)


crs(ocean_rast) <- crs(pop[[1]])
plot(ocean_rast) # 0.5 by 0.5 degree ocean raster

land_rast <- ocean_rast %>%
  mutate(name = ifelse(is.na(name), 1, NA)) 
plot(land_rast)

# ok now lets get land within 50km of the ocean
# land_dist_km <- terra::distance(ocean_rast, unit = "km")
# plot(land_dist_km)

# writeRaster(land_dist_km, here("data/model_features/ifa/int/land_dist_km.tif"))

land_dist_km <- rast(here("data/model_features/ifa/int/land_dist_km.tif"))

land_dist_50km <- land_dist_km %>%
  mutate(name = ifelse(name > 0 & name <= 50, 1, NA)) %>%
  terra::aggregate(., fact = 4, fun = sum, na.rm = TRUE) %>%
  mutate(name = ifelse(name > 0, 1, NA)) %>%
  resample(., pop[[1]], method = "near")
plot(land_dist_50km)

# now lets mask the population layer to these cells so that we only get the population of these cells

pop_50km <- mask(pop, land_dist_50km)

plot(pop_50km[[69]]) # perfect! 
global(pop_50km[[69]], "sum", na.rm = TRUE) # 1493055660
1493055660/7000000000 # 0.2132937 ~21% of the population lives within 50km of the ocean? Actually matches ok. This says 29% : https://www.preventionweb.net/news/15-global-population-lives-within-few-miles-coast-and-number-growing-rapidly


```

Calculate population density layer, final inhabited layer, and distance from inbabited cells layer and save

```{r}
area_rast <- cellSize(land_dist_50km, unit = "km")
plot(area_rast)

pop_dens <- pop_50km/area_rast
plot(pop_dens[[69]])

# Inhabited if density≥1 OR local_sum≥200; then keep only in coastal land band
inhabited <- (pop_dens >= 1) | (pop_50km >= 200) # more than 200 people within a pixel (20-60km2) or more than 1 person per km2
inhabited_fin <- terra::ifel(inhabited, 1, NA) 
writeRaster(inhabited_fin, here("data/model_features/ifa/int/inhabited_coast.tif"), overwrite = TRUE)  

plot(inhabited_fin[[69]],  main = "Inhabited") # looks good
plot(pop_dens[[69]], col = "red", main = "Population Density")

## now calculate the coastline of each inhabited cell? 

inhabited_test <- inhabited_fin[[69]]
#%>%
 # disagg(., fact = 4, method = "near") 

o <- ocean_rast %>%
  aggregate(., fact= 4)

# Ocean mask (1 = ocean, NA = land)
ocean <- resample(o, inhabited_fin, method = "near")
inhabited_dist_df_fin <- as.data.frame(ocean, xy = TRUE)

years <- 1950:2024

for(yr in 3:length(years)){
#  yr =2

# calculate distance from each inhabited cell? And then mask to ocean? 
inhabited_dist <- distance(inhabited_fin[[yr]], unit = "km") # hm it didn't like this with the whole raster stack... R crashed. 
# plot(inhabited_dist)

inhabited_dist_df <- inhabited_dist %>%
  mask(., ocean) %>%
 # resample(., data_rast) %>%
  as.data.frame(.,  xy = TRUE) %>%
  left_join(data_grid, by = c("x" = "lon", "y" = "lat"))

 #%>%
  #mutate(cutoff = 50 + 15*abs(cos(y*pi/180)))

inhabited_dist_df_fin <- full_join(inhabited_dist_df, inhabited_dist_df_fin)


}

inhabited_dist_data <- inhabited_dist_df_fin %>%
   mutate(cutoff = 50 + 15*abs(cos(y*pi/180))) %>%
  dplyr::select(-name) %>%
   pivot_longer(
    cols = matches("^\\d{4}$"),   # all columns that look like 4-digit years
    names_to = "year",
    values_to = "dist_km"
  ) %>%
  mutate(year = as.integer(year)) 

qsave(inhabited_dist_data, here("data/model_features/ifa/int/inhabited_ocean_dist.qs"))

# # test it out
# lat <- init(d_to_inhab_km, "y")  # latitude per cell
# cutoff <- 50 + 15 * abs(cos(lat * pi/180))  # add ~15 km buffer near equator
# 
# inhabited_dist_ocean <- d_to_inhab_km <= cutoff
# plot(inhabited_dist_ocean)


```


```{r}
inhabited_dist_data <- qs::qread(here("data/model_features/ifa/int/inhabited_ocean_dist.qs")) %>%
  rename(lon = x, lat = y) %>%
  left_join(data_grid) 

gfw_vars <- read.csv(here("data/model_features/gfw_static_spatial_measures.csv")) %>%
  dplyr::select(pixel_id, lon, lat, elevation_m, distance_from_shore_m) # distance to shore is actually km, not m.


ifa_fin <- inhabited_dist_data %>% 
  left_join(gfw_vars) %>%
  mutate(ifa = case_when(
    elevation_m >= -200  & dist_km <= cutoff ~ 1, 
    TRUE ~ 0
  )) %>%
  dplyr::select(pixel_id, lon, lat, year, ifa) %>%
filter(ifa == 1)

test <- ifa_fin %>%
  filter(!is.na(pixel_id),
         year == 1950) %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") # ok there are some NAs in pixel ID but they are all 0 for ifa so it doesn't matter. 

plot(test)

qs::qsave(ifa_fin, here("data/model_features/ifa/ifa_simple.qs"))

```

Now we need to gapfill any missing regions

```{r}

## now lets make sure that all of the EEZs are represented with an IFA. 
eez_lookup <- read.csv(here("data/model_features/eez/eez_lookup_fix.csv")) %>%
  clean_names() %>%
  mutate(eez_country_name = countrycode(eez_sovereign, origin = "iso3c", destination = "country.name")) %>%
  mutate(eez_country_name = case_when(
    eez_sovereign == "RAA" ~ "Azores", 
    eez_sovereign == "RAM" ~ "Madeira",
    TRUE ~ eez_country_name
  )) %>%
  mutate(eez_country_name = ifelse(is.na(eez_country_name), "High seas", eez_country_name))

eez <- read.csv(here("data/model_features/eez/eez_fix.csv"))

test <- ifa_fin %>% 
  filter(year == 1990 ) %>%
  left_join(data_grid) %>%
  left_join(eez) %>%
  left_join(eez_lookup) %>%
  # filter(eez_sovereign == "WLF") %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") %>%
  plot()

ifa_eez <- ifa_fin %>% 
  left_join(data_grid) %>%
  left_join(eez) %>%
  left_join(eez_lookup) %>%
  filter(ifa == 1)

missing_eez <- unique(setdiff(unique(eez_lookup$eez_sovereign), unique(ifa_eez$eez_sovereign)))

sort(setdiff(unique(eez_lookup$eez_country_name), unique(ifa_eez$eez_country_name)))

## we are missing these from the IFA... ? Let's see if we can fix that. I assume it has something to do with the resolution of our population layer. These all have really small EEZs
#  [1] "Algeria"                          "Antigua & Barbuda"                "Azerbaijan"                       "Azores"                          
#  [5] "Barbados"                         "Belize"                           "Bermuda"                          "Bosnia & Herzegovina"            
#  [9] "Cape Verde"                       "Comoros"                          "Cook Islands"                     "Curaçao"                         
# [13] "Djibouti"                         "Dominica"                         "French Polynesia"                 "Grenada"                         
# [17] "Guadeloupe"                       "Guam"                             "Jordan"                           "Kazakhstan"                      
# [21] "Kiribati"                         "Madeira"                          "Maldives"                         "Malta"                           
# [25] "Marshall Islands"                 "Mauritius"                        "Mayotte"                          "Micronesia (Federated States of)"
# [29] "Monaco"                           "Nauru"                            "Niue"                             "Palau"                           
# [33] "Palestinian Territories"          "Puerto Rico"                      "Réunion"                          "Samoa"                           
# [37] "São Tomé & Príncipe"              "Seychelles"                       "St. Kitts & Nevis"                "St. Lucia"                       
# [41] "St. Vincent & Grenadines"         "Syria"                            "Tonga"                            "Turkmenistan"                    
# [45] "Tuvalu"                           "Wallis & Futuna"  


ifa_fix_missing <- gfw_vars %>%
    left_join(data_grid) %>%
  left_join(eez) %>%
  left_join(eez_lookup) %>%
  filter(eez_sovereign %in% c(missing_eez))

test <- ifa_fix_missing %>%
  dplyr::select(lon, lat, elevation_m) %>%
  rast(., type = "xyz") %>%
  plot() # ok, lets just exclude the population layer from these places and only filter for <200m or <50km. Not doing AND for this one, as these are getting filtered out for some reason


  ifa_fix_missing_fin <- ifa_fix_missing %>%
      left_join(gfw_vars) %>%
  mutate(ifa = case_when(
     !(eez_country_name %in% c("Azerbaijan", "Turkmenistan", "Kazakhstan")) & (elevation_m >= -200 | distance_from_shore_m <= 50) ~ 1, 
    eez_country_name %in% c("Azerbaijan", "Turkmenistan", "Kazakhstan")  &  (elevation_m >= -200 & distance_from_shore_m <= 50) ~ 1, # caspian sea was getting completely filled in for some reason
    TRUE ~ 0
  )) 
  
  
test <- ifa_fix_missing_fin %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") %>%
  plot() 

test <- ifa_fix_missing_fin %>%
  filter(ifa == 1)

setdiff(missing_eez, unique(test$eez_sovereign)) # still missing MCO, PSE, and BIH. Are these actually in the Rousseau data? 

hist_fish_data <- qs::qread(here("data/model_features/rousseau_total_art_vessels.qs")) %>%
  mutate(flag_country_name = countrycode(flag_fin, origin = "iso3c", destination = "country.name")) %>%
  mutate(flag_country_name = 
           case_when(
             flag_fin == "RAA" ~ "Azores", 
             flag_fin == "RAM"  ~ "Madeira",
             TRUE ~ flag_country_name
           ))

  
flags <- as.data.frame(unique(hist_fish_data$flag_fin)) # get the flags we need to run models for # 151 of them

## ok cool, those missing are not actually in the Rousseau data, so we should be ok! 


ifa_fix_missing_fin <- ifa_fix_missing_fin %>%
    #  left_join(gfw_vars) %>%
  # mutate(ifa = case_when(
  #   (elevation_m >= -200 | distance_from_shore_m <= 50) ~ 1, 
  #   TRUE ~ 0
  # )) %>%
  dplyr::select(pixel_id, lon, lat, ifa) %>%
  crossing(., year = 1950:2024) %>%
  filter(ifa == 1)

test <- ifa_fix_missing_fin %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") %>%
  plot() 
  
## now add these back into the ifa 
ifa_fin_fin <- ifa_fin %>%
  rbind(., ifa_fix_missing_fin)

test <- ifa_fin_fin %>%
  filter(year == 2017) %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") %>%
  plot() 

test <- ifa_fin %>%
  filter(year == 2017) %>%
  dplyr::select(lon, lat, ifa) %>%
  rast(., type = "xyz") %>%
  plot()


qs::qsave(ifa_fin_fin, here("data/model_features/ifa/ifa_fin.qs"))

```

