---
title: "Make figures for manuscript"
output: html_document
date: "2025-10-10"
---

# Summary

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(terra)
library(foreach)
library(doParallel)
library(glue)
library(qs)
library(here)
library(countrycode)
library(strex)
library(data.table)
library(scales)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(viridis)
library(patchwork)


source(here("R/dir.R"))


data_grid <- data.table::fread(here::here("data/model_features/deg_1_x_1/global_grid.csv")) %>%
  st_drop_geometry %>%
  dplyr::select(lon, lat, pixel_id)
```

## Figure 2. Overall map figure

Plot showing 2017 effort split by artisanal and industrial 

 - make the map like this one from Kroodsma 2018: https://www.science.org/doi/10.1126/science.aao5646
   - have panel A be top left big panel of all fishing effort, b trawlers (mid and bottom), c longline, d purse seine, e all other gears, f artisanal fishing 

```{r}

# ----------------------------
# 1) Read + filter to 2017
# ----------------------------
ind_all <- fread("/home/ubuntu/data_storage/prep/random_forest/zenodo_data/mapped_industrial_effort_predictions_1950_2017.csv") %>%
  filter(year == 2017)

art_all <- fread("/home/ubuntu/data_storage/prep/random_forest/zenodo_data/artisanal_skylight_sentinel/mapped_artisanal_effort_predictions_1950_2017.csv") %>%
  filter(year == 2016)

sum(art_all$nom_active_fishing_hours)

length(unique(list.files("/home/ubuntu/data_storage/prep/random_forest/zenodo_data/artisanal_skylight_sentinel/mapped_by_flag_country")))

length(unique(list.files("/home/ubuntu/data_storage/prep/random_forest/zenodo_data/mapped_by_flag_country")))

# Land polygons
land <- ne_countries(scale = "medium", returnclass = "sf")

# ----------------------------
# 2) Your bins + colors (unchanged)
# ----------------------------
breaks_num <- c(0.02, 0.2, 2, 20, 200, 2000)
labels     <- c("<0.02", "0.02-0.2", "0.2-2", "2-20", "20-200", "200-2000", ">2000")
colors     <- c("#FFFFE4", "#EFF3FE", "#CADBEE", "#A8C9E0", "#E8F4A2", "#F1B16D", "#C54B53")
names(colors) <- labels

# ----------------------------
# 3) Helper to summarise -> kW hours/km2 -> bin
# ----------------------------
summarise_eff <- function(df, title, gear_filter = NULL) {
  d <- df
  if (!is.null(gear_filter)) {
    d <- gear_filter(d)
  }

  area <- d %>% distinct(lon, lat, pixel_area_km2)

  out <- d %>%
    group_by(lon, lat) %>%
    summarise(nom_eff = sum(nom_active_fishing_hours, na.rm = TRUE), .groups = "drop") %>%
    filter(nom_eff > 0) %>%
    left_join(area, by = c("lon", "lat")) %>%
    mutate(
      eff_km2 = nom_eff / pixel_area_km2,
      eff_bin = cut(
        eff_km2,
        breaks = c(-Inf, breaks_num, Inf),
        labels = labels,
        right = FALSE,
        include.lowest = TRUE
      ),
      eff_bin = factor(eff_bin, levels = labels, ordered = TRUE),
      panel = title
    )

  out
}

# ----------------------------
# 4) Build panel datasets (A–F)
# ----------------------------
pA <- summarise_eff(ind_all, "A.  Industrial (Total)")

pB <- summarise_eff(
  ind_all, "B.  Industrial (Trawlers)",
  gear_filter = function(d) d %>% filter(gear %in% c("Trawl_Bottom", "Trawl_Midwater_or_Unsp"))
)

pC <- summarise_eff(
  ind_all, "C.  Industrial (Longlines)",
  gear_filter = function(d) d %>% filter(gear == "Lines_Longlines")
)

pD <- summarise_eff(
  ind_all, "D.  Industrial (Purse Seines)",
  gear_filter = function(d) d %>% filter(gear == "Seine_Purse_Seine")
)

pE <- summarise_eff(
  ind_all, "E.  Industrial (Other gears)",
  gear_filter = function(d) d %>%
    filter(!(gear %in% c("Seine_Purse_Seine", "Lines_Longlines", "Trawl_Bottom", "Trawl_Midwater_or_Unsp")))
)

pF <- summarise_eff(art_all, "F.  Artisanal (Total)")

# Common extent across all panels
pall <- bind_rows(pA, pB, pC, pD, pE, pF)

xlims <- range(pall$lon, na.rm = TRUE)
ylims <- range(pall$lat, na.rm = TRUE)

# ----------------------------
# 6) Patchwork layout like the example
#    A spans top-left (2 cols x 2 rows)
#    B/C/D stacked on the right
#    Bottom row: F, E, D
# ----------------------------
# 1) A tighter, paper-style theme for maps
theme_map_paper <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      # remove grid + all axis stuff
      panel.grid      = element_blank(),
      axis.title      = element_blank(),
      axis.text       = element_blank(),
      axis.ticks      = element_blank(),

      # tighten spacing / margins
      plot.margin     = margin(2, 2, 2, 2),
      panel.spacing   = unit(2, "mm"),

      # clean borders
      panel.border    = element_rect(colour = "grey30", fill = NA, linewidth = 0.35),

      # titles like the example (left aligned, bold)
      plot.title      = element_text(face = "bold", size = 14, hjust = 0),

      # legend
      legend.position = "bottom",
      legend.title    = element_text(face = "bold"),
      legend.text     = element_text(size = 10),

      # make legend keys a bit larger / more readable
      legend.key.height = unit(5, "mm"),
      legend.key.width  = unit(10, "mm")
    )
}

# 2) Replace your panel_plot() with this version (same inputs)
panel_plot <- function(pdat, show_legend = FALSE) {
  ggplot() +
    geom_raster(data = pdat, aes(lon, lat, fill = eff_bin)) +
    geom_sf(data = land, fill = "grey85", color = "grey60", linewidth = 0.15) +
    coord_sf(
      crs = st_crs(4326),
      xlim = xlims,
      ylim = ylims,
      expand = FALSE
    ) +
    scale_fill_manual(
      values = colors,
      drop = FALSE,
      name = "Fishing Effort (kW hours/km²)",
      guide = guide_legend(
        nrow = 1,
        byrow = TRUE,
        title.position = "top",
        title.hjust = 0.5
      )
    ) +
    labs(title = unique(pdat$panel)) +
    theme_map_paper() +
    theme(
      # only show legend on one panel (we'll collect later)
      legend.position = if (show_legend) "bottom" else "none"
    )
}

# 3) Rebuild panels (same as before, but now with upgraded styling)
gA <- panel_plot(pA)
gB <- panel_plot(pB)
gC <- panel_plot(pC)
gD <- panel_plot(pD, show_legend = TRUE)  # keep legend here so it gets collected
gE <- panel_plot(pE)
gF <- panel_plot(pF)

# 4) Patchwork layout + collect legend + tighten whitespace a bit more
design <- "
AAB
AAC
DEF
"

p_final <- (gA + gB + gC + gF + gE + gD) +
  plot_layout(design = design, guides = "collect") &
  theme(
    legend.position = "bottom",
    # extra tightening at the combined-plot level
    plot.margin   = margin(2, 2, 2, 2)
  )

# p_final

ggsave(
  here("prep/06_manuscript/figs/fig2_industrial_by_gear_plus_artisanal_2017.png"),
  plot = p_final,
  width = 14, height = 8, dpi = 600, bg = "white"
)

```

## SI Figure 3

Make a zoomed in map of some artisanal places

```{r}

summary(art_all)


zooms <- list(
  "A. West Africa"   = list(x = c(-20,  20), y = c(-5,  25)),
  "B. Indian Peninsula"         = list(x = c( 66,  95), y = c(  5,  26)),
  "C. Indonesian Archipelago"     = list(x = c( 95, 125), y = c(-12,   8)),
  "D. East Asia" = list(x = c(105, 148), y = c( 18,  47))
)

make_zoom_panel <- function(region_name, bbox, pdat, land, fill_var = "eff_bin",
                            colors = NULL, fill_name = "Fishing effort") {

  ggplot() +
    geom_raster(
      data = pdat,
      aes(x = lon, y = lat, fill = .data[[fill_var]])
    ) +
    geom_sf(data = land, fill = "grey85", color = "grey60", linewidth = 0.1) +
    coord_sf(
      xlim = bbox$x, ylim = bbox$y,
      expand = FALSE
    ) +
    labs(title = region_name, x = NULL, y = NULL, fill = fill_name) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid = element_blank(),
        axis.ticks = element_blank(),
  axis.text  = element_blank()
    ) +
    { if (!is.null(colors)) scale_fill_manual(values = colors, drop = FALSE) else NULL }
}



p_list <- lapply(names(zooms), \(nm) {
  make_zoom_panel(
    region_name = nm,
    bbox = zooms[[nm]],
    pdat = pF,
    land = land,
    fill_var = "eff_bin",              # <- change if needed
    colors = colors,                   # <- your existing palette (optional)
    fill_name = "Fishing Effort (kW hours/km²)"  # <- match main figure
  )
})


p_combined <-
  (p_list[[1]] | p_list[[2]]) /
  (p_list[[3]] | p_list[[4]]) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

p_combined

# Save the 4-panel patchwork figure as a PNG
out_dir <- "prep/06_manuscript/figs"

out_file <- file.path(here(out_dir), "SI_fig_artisanal_zooms.png")

ggsave(
  filename = out_file,
  plot     = p_combined,
  width  = 180,   # mm (≈ 7.1 in)
  height = 180,   # mm
  units  = "mm",
  dpi    = 600,
  bg     = "white"
)


```



## Figure 3
Plot showing 2017 maps comparing our results, AIS from GFW, and Rousseau et al. 2024

```{r}
options(timeout = max(3600, getOption("timeout")))  # e.g. 1 hour
rousseau_ind_2017 <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_2017_I.csv") %>%
  dplyr::select(-V1)

cell_area_rousseau <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/CellstoLME_EEZ.csv") %>%
  distinct(lon = Lon, lat = Lat, pixel_area_km2 = Area)

convert_fishing_hours <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/ConversionFishingHours.csv") %>%
  dplyr::select(-V1)

rousseau_ind_2017_hours <- rousseau_ind_2017 %>%
  left_join(convert_fishing_hours) %>%
  mutate(NomActiveHours = NomActive*mean)

rousseau_2017 <- rousseau_ind_2017_hours %>%
  mutate(year = 2017) %>%
  group_by(lon = Lon, lat = Lat) %>%
  summarise(nom_eff = sum(NomActiveHours, na.rm = TRUE)) %>%
  ungroup() %>%
  na.omit() %>%
  filter(nom_eff > 0) %>%
  transmute(lon, lat, nom_eff, source = "Rousseau et al 2024") %>%
  left_join(cell_area_rousseau) %>%
  mutate(eff_km2 = nom_eff/pixel_area_km2)

new_data <- fread("/home/ubuntu/data_storage/prep/random_forest/zenodo_data/mapped_industrial_effort_predictions_1950_2017.csv") %>%
  filter(year == 2017)

area_new <- new_data %>%
  distinct(lon, lat, pixel_area_km2)

new_ind <- new_data %>%
  group_by(lon, lat) %>%
  summarise(nom_eff = sum(nom_active_fishing_hours, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(nom_eff > 0) %>%
  transmute(lon, lat, nom_eff, source = "Model") %>%
  left_join(area_new) %>%
  mutate(eff_km2 = nom_eff/pixel_area_km2)
sum(new_ind$nom_eff) # 114873183573


gfw_vessel_registry <- qread(here("prep/06_manuscript/data/gfw_vessel_registry_fix.qs")) %>%
  dplyr::distinct(mmsi, flag_gfw, length_category, gear, tonnage_gt_gfw, engine_power_kw_gfw)

gfw_2017 <- qread(file.path(rdsi_raw_dir, "global_fishing_watch/apparent_fishing_hours_mmsi/v3_aggregated/all_effort_2017.qs")) %>%
  dplyr::select(x = cell_ll_lon, y = cell_ll_lat,  mmsi, hours, fishing_hours) %>%
  left_join(gfw_vessel_registry)
  

# gfw_2017 <- qs::qread("/home/ubuntu/data_storage/prep/gfw_props/deg_half/all_effort_gear_length_props_2017.qs")

res <- 0.1
deg_to_km <- 111.32

gfw_2017_area <- gfw_2017 %>%
  mutate(pixel_area_km2 = (res * deg_to_km) * (res * deg_to_km * cos(pi * y / 180))) %>%
  distinct(lon = x, lat = y, pixel_area_km2)


gfw_data <- gfw_2017 %>%
  mutate(kw_fishing_hours = fishing_hours * engine_power_kw_gfw) %>%
    group_by(lon = x, lat = y) %>%
  summarise(nom_eff = sum(kw_fishing_hours, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(nom_eff > 0) %>%
  transmute(lon, lat, nom_eff, source = "AIS (GFW)") %>%
  left_join(gfw_2017_area) %>%
  mutate(eff_km2 = nom_eff/pixel_area_km2)



# Your custom bins
breaks_num <- c(0.02, 0.2, 2, 20, 200, 2000)
labels     <- c("<0.02", "0.02-0.2", "0.2-2", "2-20", "20-200", "200-2000", ">2000")
colors     <- c("#FFFFE4", "#EFF3FE", "#CADBEE", "#A8C9E0", "#E8F4A2", "#F1B16D", "#C54B53")
names(colors) <- labels

# Combine old + new
pdat <- bind_rows(gfw_data, new_ind, rousseau_2017) %>%
  mutate(
    eff_bin = cut(
      eff_km2,
      breaks = c(-Inf, breaks_num, Inf),   # add open-ended bins
      labels = labels,
      right = FALSE,
      include.lowest = TRUE
    ),
    eff_bin = factor(eff_bin, levels = labels, ordered = TRUE)
  ) %>%
    mutate(
    source = factor(source, levels = c("AIS (GFW)", "Model" , "Rousseau et al 2024"))
  )



# 1. Get global land polygons (at medium resolution)
land <- ne_countries(scale = "medium", returnclass = "sf")

# Plot with coord_sf()
p <- ggplot() +
  geom_raster(data = pdat, mapping = aes(lon, lat, fill = eff_bin)) +
  geom_sf(data = land, fill = "grey85", color = "grey60", linewidth = 0.1) +
  facet_wrap(~ source, nrow = 3) +
  coord_sf(
    xlim = range(pdat$lon, na.rm = TRUE),
    ylim = range(pdat$lat, na.rm = TRUE),
    expand = FALSE
  ) +
  scale_fill_manual(
    values = colors,
    drop = FALSE,
    name = "Fishing Effort (kW hours/km²)",
    guide = guide_legend(
      nrow = 1,
      byrow = TRUE,
      title.position = "top",
      title.hjust = 0.5
    )
  ) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )

p

# Make sure the facet order is what you want
pdat$source <- factor(pdat$source, levels = sort(unique(pdat$source)))

labs_vec <- setNames(
  paste0(LETTERS[seq_along(levels(pdat$source))], ". ", levels(pdat$source)),
  levels(pdat$source)
)

p_labeled <- p +
  facet_wrap(~ source, nrow = 3, labeller = labeller(source = labs_vec)) +
  theme(strip.text = element_text(hjust = 0)  # 0 = left, 0.5 = center, 1 = right
        )

p_labeled

ggsave(
  here("prep/06_manuscript/figs/fig3_fishing_effort_comp_2017.png"),
  plot   = p_labeled,
  width  = 8,   # a bit less width
  height = 10,  # more height for the 3 maps
  dpi    = 600,
  bg     = "white"
)


```

## SI Figure 4

Make a plot of effort binned by lat and lon, split by length category comparing Rousseau et al and our data, similar to figure 8 from Rousseau et al. 2024: https://www.nature.com/articles/s41597-023-02824-6#Sec13

```{r}
mapped_2017_apw <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_2017_APW.csv")

mapped_2017_up <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_2017_UP.csv")

all_rousseau_2017_art <- # rbind(mapped_2017_apw, mapped_2017_up) %>%
  mapped_2017_apw %>%
  left_join(convert_fishing_hours) %>%
  mutate(NomActiveHours = NomActive*mean) %>%
  dplyr::select(-V1)


round_to_one <- function(x) {
  round(x)
}

rousseau_1_deg <- rousseau_ind_2017_hours %>%
  rbind(all_rousseau_2017_art) %>%
    group_by(lon = Lon, lat = Lat, length_category = Length_Category) %>%
  summarise(nom_eff = sum(NomActiveHours, na.rm = TRUE)) %>%
  ungroup() %>%
    mutate(
    x_1 = round_to_one(lon),
    y_1 = round_to_one(lat)
  ) %>%
  # Group by rounded coordinates and existing grouping variables
  group_by(x_1, y_1, length_category) %>%
  # Aggregate numeric columns
  summarise(
    nom_eff = sum(nom_eff, na.rm = TRUE) # Sum the vessel counts
  ) %>%
  ungroup() %>%
  # Rename coordinates back to original names if needed
  rename(lon = x_1, lat = y_1) %>%
  as.data.table() %>%
  left_join(data_grid)

# gapfill missing pixels 
# split into known vs missing pixel_id
known <- rousseau_1_deg %>% filter(!is.na(pixel_id))
miss  <- rousseau_1_deg %>% filter(is.na(pixel_id))

# make sf points and project to a metric CRS for nearest-neighbor distance
known_sf <- st_as_sf(known, coords = c("lon", "lat"), crs = 4326) %>% st_transform(3857)
miss_sf  <- st_as_sf(miss,  coords = c("lon", "lat"), crs = 4326) %>% st_transform(3857)

# for each missing point, find index of nearest known point
nn_idx <- st_nearest_feature(miss_sf, known_sf)

# reassign missing pixel_ids
rousseau_1_deg$pixel_id[is.na(rousseau_1_deg$pixel_id)] <- known$pixel_id[nn_idx]

rousseau_1_deg <- rousseau_1_deg %>%
  group_by(pixel_id, length_category) %>%
  summarise(nom_eff = sum(nom_eff, na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(data_grid) %>%
  filter(!is.na(pixel_id)) %>%
  mutate(source = "Rousseau et al. 2024")


## Now join with our 2017 effort
new_data_art <- fread("/home/ubuntu/data_storage/prep/random_forest/zenodo_data/artisanal_skylight_sentinel/mapped_artisanal_effort_predictions_1950_2017.csv") %>%
  filter(year == 2017) %>%
  dplyr::select(colnames(new_data %>% dplyr::select(-gear)))

all_eff <- new_data %>%
  dplyr::select(-gear) %>%
  rbind(new_data_art) %>%
  group_by(lon, lat, length_category) %>%
  summarise(nom_eff = sum(nom_active_fishing_hours, na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(data_grid) %>%
    mutate(source = "Model") %>%
  rbind(., rousseau_1_deg)


lat_prof <- all_eff %>%
  group_by(source, lat, length_category) %>%
  summarise(effort = sum(nom_eff, na.rm = TRUE), .groups = "drop")

lon_prof <- all_eff %>%
  group_by(source, lon, length_category) %>%
  summarise(effort = sum(nom_eff, na.rm = TRUE), .groups = "drop")


cols <- c("Model" = "black", "Rousseau et al. 2024" = "red")

p_lat <- ggplot(lat_prof, aes(x = lat, y = effort, color = source)) +
  geom_line(linewidth = 0.5) +
  scale_color_manual(values = cols) +
  labs(x = "Latitude", y = "Effort") +
  theme_classic() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.8, 0.15),
    legend.title = element_blank()
  ) +
  facet_wrap(~ length_category, nrow = 3, scales = "free")

p_lon <- ggplot(lon_prof, aes(x = lon, y = effort, color = source)) +
  geom_line(linewidth = 0.5) +
  scale_color_manual(values = cols) +
  labs(x = "Longitude", y = "Effort") +
  theme_classic() +
  theme(
    legend.position = "none"
  ) +
  facet_wrap(~ length_category, nrow = 3, scales = "free")

full_plot <- (p_lat / p_lon)

ggsave(
  here("prep/06_manuscript/figs/SI_fig_comnpare_lat_lon.png"),
  plot   = full_plot,
  width  = 9,   # a bit less width
  height = 11,  # more height for the 3 maps
  dpi    = 600,
  bg     = "white"
)

lon_prof %>% 
  filter(source == "Model") %>%
  pull(effort) %>%
  sum()

lon_prof %>% 
  filter(source == "Rousseau et al. 2024") %>%
  pull(effort) %>%
  sum()


test <- fread("/home/ubuntu/data_storage/prep/random_forest/zenodo_data/artisanal_skylight_sentinel/known_artisanal_effort_rousseau.csv") %>%
  filter(year == 2017)

sum(test$modeled_nom_active_fishing_days)
sum(test$total_nom_days) # ok so we aren't missing any here... not sure why artisanal is different when loading spatialized rousseau data.
```

## SI Figure 7
Lets make a plot showing effort in China in 1950 vs 1985 vs 2017 for industrial 

```{r}
cell_area_rousseau <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/CellstoLME_EEZ.csv") %>%
  distinct(lon = Lon, lat = Lat, pixel_area_km2 = Area)

rousseau_ind_2017 <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_2017_I.csv")
rousseau_ind_1985 <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_1985_I.csv")
rousseau_ind_1950 <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_1950_I.csv")

rousseau_ind_2017_hours <- rousseau_ind_2017 %>%
  left_join(convert_fishing_hours) %>%
  mutate(NomActiveHours = NomActive*mean)

rousseau_ind_1985_hours <- rousseau_ind_1985 %>%
  left_join(convert_fishing_hours) %>%
  mutate(NomActiveHours = NomActive*mean)


rousseau_ind_1950_hours <- rousseau_ind_1950 %>%
  left_join(convert_fishing_hours) %>%
  mutate(NomActiveHours = NomActive*mean)


rousseau_chn <- rousseau_ind_2017_hours %>%
  mutate(year = 2017) %>%
  filter(SAUP == 156) %>%
  rbind(rousseau_ind_1985_hours %>% filter(SAUP == 156) %>% mutate(year = 1985)) %>%
  rbind(rousseau_ind_1950_hours %>% filter(SAUP == 156) %>% mutate(year = 1950)) %>%
  group_by(year, lon = Lon, lat = Lat) %>%
  summarise(nom_eff = sum(NomActiveHours, na.rm = TRUE)) %>%
  ungroup() %>%
  na.omit() %>%
  filter(nom_eff > 0) %>%
  transmute(year, lon, lat, nom_eff, source = "Rousseau et al 2024") %>%
  left_join(cell_area_rousseau) %>%
  mutate(eff_km2 = nom_eff/pixel_area_km2)
  
# SAUP = 156 for China
area_new <- fread(file.path(rdsi_dir, "prep/random_forest/zenodo_data/mapped_by_flag_country/model_preds_1950_2017_CHN.csv")) %>%
  distinct(lon, lat, pixel_area_km2)

new_china_ind <- fread(file.path(rdsi_dir, "prep/random_forest/zenodo_data/mapped_by_flag_country/model_preds_1950_2017_CHN.csv")) %>%
  filter(year %in% c(1950, 1985, 2017)) %>%
  group_by(year, lon, lat) %>%
  summarise(nom_eff = sum(nom_active_fishing_hours, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(nom_eff > 0) %>%
  transmute(year, lon, lat, nom_eff, source = "Model") %>%
  left_join(area_new) %>%
  mutate(eff_km2 = nom_eff/pixel_area_km2)

years <- c(1950, 1985, 2017)

pdat <- bind_rows(new_china_ind, rousseau_chn) %>%
  mutate(
    year   = factor(year, levels = years),
    source = factor(source, levels = c("Model", "Rousseau et al 2024"))
  )

# consistent extents across all facets
lims <- pdat %>%
  summarise(xmin = min(lon), xmax = max(lon),
            ymin = min(lat), ymax = max(lat))



# Your custom bins
breaks_num <- c(0.02, 0.2, 2, 20, 200, 2000)
labels     <- c("<0.02", "0.02-0.2", "0.2-2", "2-20", "20-200", "200-2000", ">2000")
colors     <- c("#FFFFE4", "#EFF3FE", "#CADBEE", "#A8C9E0", "#E8F4A2", "#F1B16D", "#C54B53")
names(colors) <- labels

# Combine old + new
pdat <- bind_rows(rousseau_chn, new_china_ind) %>%
  mutate(
    eff_bin = cut(
      eff_km2,
      breaks = c(-Inf, breaks_num, Inf),   # add open-ended bins
      labels = labels,
      right = FALSE,
      include.lowest = TRUE
    ),
    eff_bin = factor(eff_bin, levels = labels, ordered = TRUE)
  )


# 1. Get global land polygons (at medium resolution)
land <- ne_countries(scale = "medium", returnclass = "sf")

# Plot with coord_sf()
(p <- ggplot() +
  geom_raster(data = pdat, aes(lon, lat, fill = eff_bin)) +
  geom_sf(data = land, fill = "grey85", color = "grey60", linewidth = 0.1) +
  facet_grid(year ~ source) +
  coord_sf(
    xlim = range(pdat$lon, na.rm = TRUE),
    ylim = range(pdat$lat, na.rm = TRUE),
    expand = FALSE
  ) +
  scale_fill_manual(
    values = colors,
    drop = FALSE,
    name = "Fishing Effort (kW hours/km²)",
    guide = guide_legend(
      nrow = 1,
      byrow = TRUE,
      title.position = "top",
      title.hjust = 0.5
    )
  ) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  ))


# 4. Save
ggsave(here("prep/06_manuscript/figs/SI_fig5_fishing_effort_china_with_land.png"), plot = p, width = 10, height = 6, dpi = 600, bg = "white")



```


## SI Figure 2
Inshore fishing areas 

```{r}

ifa_fin_fin <- qs::qread(here("data/model_features/ifa/ifa_fin.qs"))

# cool! 

test_df <- ifa_fin_fin %>% 
  filter(year == 2017) %>%
  select(lon, lat, ifa) %>%
  mutate(ifa = factor(ifa, levels = c(0, 1), labels = c("No IFA", "IFA")))

ggplot(test_df) +
  geom_raster(aes(x = lon, y = lat, fill = ifa)) +
  coord_equal() +
  scale_fill_manual(
    name = "IFA status",
    values = c("No IFA" = "grey85", "IFA" = "#1b9e77"),
    drop = FALSE
  ) +
  labs(
    x = "Inshore fishing areas",
    y = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text = element_text(size = 10),
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  )

ggsave(here("figs/SI_fig2_ifa_layer.png"))
```

## SI Figure 1
Sea ice

```{r}

test <- qs::qread(here("data/model_features/sea_ice_features_lit.qs"))

test_rast <- test %>%
  filter(year == 1950) %>%
  dplyr::select(lon, lat, n_months_ice) 

# Convert raster to dataframe for ggplot
# test_df <- as.data.frame(test_rast, xy = TRUE, na.rm = TRUE)

# Plot
ggplot(test_rast) +
  geom_raster(aes(x = lon, y = lat, fill = n_months_ice)) +
  scale_fill_viridis(name = "", option = "C", na.value = "grey90") +
  coord_equal() +
  labs(
    x = "Months with ice coverage",
    y = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text = element_text(size = 10),
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

    ggsave(here("figs/SI_fig1_ice_layer.png"))

```

## Stuff for Figure 1
Simple line chart of effort hours over time for schematic 

```{r}
rousseau_effort <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/CapacityCountryLevel_Detailed.csv")

over_time <- rousseau_effort %>%
  group_by(Year) %>%
  summarise(total_hours = sum(NomActive, na.rm = TRUE)) %>%
  ungroup()


ggplot(over_time, aes(x = Year, y = total_hours)) +
  geom_line(linewidth = 1) +
  labs(
    x     = NULL,
    y     = NULL
  ) +
  scale_x_continuous(
    breaks = c(1950, 1960, 1970, 1980, 1990, 2000, 2010),
    limits = c(1950, 2017),
   # expand = c(0, 0)
  ) +
  theme_classic(base_size = 32) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.title.y  = element_text(margin = margin(r = 8)),
    axis.line.x   = element_line(),
    axis.line.y   = element_line(),
    axis.text.y   = element_blank(),
    axis.ticks.y  = element_blank(),
    panel.border  = element_blank())
  

```


Make map showing summed GFW effort from 2015-2024 to add to the flowchart

```{r}
gfw_vessel_registry <- qread(here("prep/06_manuscript/data/gfw_vessel_registry_fix.qs")) %>%
  dplyr::distinct(mmsi, flag_gfw, length_category, gear, tonnage_gt_gfw, engine_power_kw_gfw)

gfw_files <- list.files(file.path(rdsi_raw_dir, "global_fishing_watch/apparent_fishing_hours_mmsi/v3_aggregated"), full.names = TRUE)[13]

gfw_all <- lapply(gfw_files, qread) %>%
  bind_rows() %>%
  dplyr::select(x = cell_ll_lon, y = cell_ll_lat,  mmsi, hours, fishing_hours) %>%
  left_join(gfw_vessel_registry)


res <- 0.1
deg_to_km <- 111.32

gfw_all_area <- gfw_all %>%
  mutate(pixel_area_km2 = (res * deg_to_km) * (res * deg_to_km * cos(pi * y / 180))) %>%
  distinct(lon = x, lat = y, pixel_area_km2)


gfw_data <- gfw_all %>%
  mutate(kw_fishing_hours = fishing_hours * engine_power_kw_gfw) %>%
    group_by(lon = x, lat = y) %>%
  summarise(nom_eff = sum(kw_fishing_hours, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(nom_eff > 0) %>%
  transmute(lon, lat, nom_eff, source = "AIS (GFW)") %>%
  left_join(gfw_all_area) %>%
  mutate(eff_km2 = nom_eff/pixel_area_km2)


summary(gfw_data)



# consistent extents across all facets
lims <- gfw_data %>%
  summarise(xmin = min(lon), xmax = max(lon),
            ymin = min(lat), ymax = max(lat))



# Your custom bins
breaks_num <- c(0.02, 0.2, 2, 20, 200, 2000)
labels     <- c("<0.02", "0.02-0.2", "0.2-2", "2-20", "20-200", "200-2000", ">2000")
colors     <- c("#FFFFE4", "#EFF3FE", "#CADBEE", "#A8C9E0", "#E8F4A2", "#F1B16D", "#C54B53")
names(colors) <- labels

# Combine old + new
pdat <- bind_rows(gfw_data) %>%
  mutate(
    eff_bin = cut(
      eff_km2,
      breaks = c(-Inf, breaks_num, Inf),   # add open-ended bins
      labels = labels,
      right = FALSE,
      include.lowest = TRUE
    ),
    eff_bin = factor(eff_bin, levels = labels, ordered = TRUE)
  )

library(tibble)
library(ggplot2)
library(maps)
library(mapproj)

# outline of the full world domain in lon/lat
world_frame <- tibble(
  lon = c(
    seq(-180, 180, length.out = 400),      # bottom edge
    rep(180, 400),                         # right edge
    seq(180, -180, length.out = 400),      # top edge
    rep(-180, 400)                         # left edge
  ),
  lat = c(
    rep(-89.9, 400),
    seq(-89.9, 89.9, length.out = 400),
    rep(89.9, 400),
    seq(89.9, -89.9, length.out = 400)
  )
)

ggplot() +
  geom_raster(data = pdat, aes(lon, lat, fill = eff_bin)) +
  borders("world", fill = "grey85", colour = "grey60", size = 0.1) +
  
  # **this becomes the oval border**
  geom_path(
    data = world_frame,
    aes(lon, lat),
    inherit.aes = FALSE,
    linewidth = 0.6,
    colour = "black"
  ) +
  
  ggplot2::coord_map(
    projection = "mollweide",
    xlim = c(-180, 180),
    ylim = c(-89.9, 89.9)
  ) +
  scale_fill_manual(values = colors, drop = FALSE) +
  theme_void() +
  theme(
    legend.position  = "none",
    panel.border     = element_blank(),   # no square border
    panel.background = element_rect(fill = "white", colour = NA)
  )

```


## Exploration

Lets compare our artisanal modelling to Rousseau et al. artisanal data 

```{r}
saup_codes <- read.csv("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/SAUPtoCountry.csv")

# lets load in the mapped artisanal powered and unpowered and join
options(timeout = 600)  # extend timeout to 10 minutes

mapped_2017_apw <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_2017_APW.csv")

mapped_2017_up <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_2017_UP.csv")

all_rousseau_2017 <- rbind(mapped_2017_apw, mapped_2017_up)

  all_files_flag <- list.files(file.path(rdsi_dir, "prep/random_forest/zenodo_data/artisanal_skylight_sentinel/mapped_by_flag_country/"), full.names = TRUE)

    all_data <- lapply(all_files_flag, fread) %>%
      bind_rows()
    
    all_data_2017 <- all_data %>%
      filter(year == 2017)
    
test_new <- all_data_2017 %>%    
      group_by(lon, lat) %>%
      summarise(effort = sum(nom_active_fishing_hours, na.rm = TRUE)) %>%
      ungroup() %>%
      filter(effort > 0) %>%
      rast(., type = "xyz")
    
    
test_rousseau <- all_rousseau_2017 %>%
  group_by(Lon, Lat) %>%
  summarise(nom_eff = sum(NomActive, na.rm = TRUE)) %>%
  ungroup() %>%
  na.omit() %>%
  filter(nom_eff > 0) %>%
  rast(., type = "xyz")


par(mfrow = c(2, 1))  # 2 rows, 1 column
plot(test_new, main = "Predicted effort (new model)")
plot(test_rousseau, main = "Rousseau et al. 2019 effort")

# interesting, looks pretty similar, however, ours is definitely 1) more concentrated, and 2) closer to shore. I am a little confused because presumably Watson (and by extension Rousseau) have the same <=50km or >=200m depth as us, but it looks like their artisnal goes out to the edge of EEZs (very low effort all the way out on the edge though). Just looking at unpowered looks a lot more similar.

```

```{r}
rousseau_effort <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/CapacityCountryLevel_Detailed.csv")

ids <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/SAUPtoCountry.csv")

test <- rousseau_effort %>%
  filter(Country == "AUS") %>%
  group_by(Country, Year, Length_Category) %>%
  summarise(nom_eff = sum(NomActive, na.rm = TRUE),
            nv = sum(NV, na.rm = TRUE)) %>%
  ungroup()

test <- rousseau_effort %>%
  filter(Country == "AUS",
         Year == 2017)


test <- rousseau_effort %>%
  filter(Year == 2017,
         Country %in% c("AUS", "NZL")) %>%
  group_by(Country) %>%
  summarise(nom_eff = sum(NomActive, na.rm = TRUE)) %>%
  ungroup()

test %>%
  filter(Country == "AUS") %>%
  pull(nom_eff) # 340036707

test %>%
  filter(Country == "NZL") %>%
  pull(nom_eff) # 32911763
 
340036707/32911763 # 10.3


## look at number of vessels? 
test <- rousseau_effort %>%
  filter(Year == 2013,
         Country %in% c("AUS", "NZL")) %>%
  group_by(Country, Sector) %>%
  summarise(nom_eff = sum(NomActive, na.rm = TRUE),
            nv = sum(NV, na.rm = TRUE)) %>%
  ungroup()

## read in only effort fished in Australia and NZL EEZ? 

rousseau_ind_2017 <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_2017_I.csv") %>%
  dplyr::select(-V1)

rousseau_artp_2017 <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_2017_APW.csv") %>%
  dplyr::select(-V1)

rousseau_up_2017 <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_2017_UP.csv") %>%
  dplyr::select(-V1)

cells_to_eez <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/CellstoLME_EEZ.csv") %>%
  dplyr::select(-V1)

all_df <- rbind(rousseau_up_2017, rousseau_artp_2017, rousseau_ind_2017) %>%
  left_join(cells_to_eez)
  

aus_nzl_eez <- all_df %>% 
  filter(EEZ %in% c("Australia", "New Zealand")) %>%
  group_by(EEZ) %>%
  summarise(nom_eff = sum(NomActive, na.rm = TRUE)) %>%
  ungroup()

aus_nzl_eez %>%
  filter(EEZ == "Australia") %>%
  pull(nom_eff) # 332977954

aus_nzl_eez %>%
  filter(EEZ == "New Zealand") %>%
  pull(nom_eff) # 36967241

332977954/36967241 # 9

aus_nzl_eez_flag <- all_df %>% 
  filter(EEZ %in% c("Australia", "New Zealand")) %>%
  filter(SAUP %in% c(36, 554)) %>%
  group_by(SAUP, EEZ) %>%
  summarise(nom_eff = sum(NomActive, na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(ids)

aus_nzl_eez_flag %>%
  filter(EEZ == "Australia", 
         Country == "AUS") %>%
  pull(nom_eff) # 311574210

aus_nzl_eez_flag %>%
  filter(EEZ == "New Zealand", 
         Country == "NZL") %>%
  pull(nom_eff) # 32381806

311574210/32381806 # 9.6

```

Look at catch differences between NZL and AUS for 2017 from Watson data?

```{r}

catch <- fread("/home/ubuntu/data_storage/raw_data/fishing_effort_catch_from_gem/catch_histsoc_1869_2017_EEZ_addFAO.csv")

aus_nzl_catch <- catch %>%
  filter(Year == 2017,
         FCountryName %in% c("Australia", "New_Zealand")) %>%
  mutate(total_catch = Reported + IUU) %>%
  group_by(Year, FCountryName, Sector) %>%
  summarise(total_catch = sum(total_catch, na.rm = TRUE)) %>%
  ungroup()

```

```{r}
library(janitor)

r_eff_new <- rousseau_effort %>%
    filter(!is.na(Length_Category)) %>%
  clean_names() %>%
  group_by(flag_fin = country, length_category, year, sector) %>%
  summarise(nom_active_days = sum(nom_active, na.rm = TRUE),
            eff_active_days = sum(eff_active, na.rm = TRUE), 
            nv = sum(nv, na.rm = TRUE),
            p = sum(p),
            gt = sum(gt)) %>%
  ungroup() %>%
  filter(nv > 0) %>%
  mutate(days_at_sea = nom_active_days/p,
         mean_engine_p = p/nv) 

summary(r_eff_new)


r_eff_new <- rousseau_effort %>%
    filter(!is.na(Length_Category)) %>%
  dplyr::select(-V1) %>%
  clean_names() %>%
  filter(nv > 0,
       country %in% c("AUS", "NZL"),
       year == 2017) %>%
  mutate(days_at_sea = nom_active/p_active) 

summary(r_eff_new)


country_decomp <- r_eff_new %>%
  group_by(country) %>%
  summarise(
    nom_eff_total = sum(nom_active, na.rm = TRUE),
    nv_active_total = sum(nv_active, na.rm = TRUE),
    p_active_total  = sum(p_active,  na.rm = TRUE),

    # implied means
    mean_power_per_active_vessel = p_active_total / nv_active_total,
    mean_days_at_sea             = nom_eff_total / p_active_total,

    # this should reconstruct nom_eff_total if the identity holds
    reconstructed_nom_eff = nv_active_total *
      mean_power_per_active_vessel *
      mean_days_at_sea,

    reconstruction_ratio = reconstructed_nom_eff / nom_eff_total,
    .groups = "drop"
  )

country_decomp

# AUS has a much larger active fleet (~6500 vessels in AUS vs 882 vessels in NZL)
# AUS has a larger mean power per active vessel (258.3635) than NZL (186.6633)
# AUS has a slightly higher mean days at sea (201.7238) than NZL (199)

ratio_table <- country_decomp %>%
  select(country, nom_eff_total, nv_active_total,
         mean_power_per_active_vessel, mean_days_at_sea) %>%
  tidyr::pivot_wider(names_from = country, values_from = everything()) %>%
  mutate(
    ratio_nom_eff = nom_eff_total_AUS / nom_eff_total_NZL,
    ratio_nv      = nv_active_total_AUS / nv_active_total_NZL,
    ratio_power   = mean_power_per_active_vessel_AUS / mean_power_per_active_vessel_NZL,
    ratio_das     = mean_days_at_sea_AUS / mean_days_at_sea_NZL,

    ratio_product = ratio_nv * ratio_power * ratio_das
  )

ratio_table

# ratio_nv = 7.390663
# ratio_power = 1.384115
# ratio_das = 1.009995

gear_totals <- r_eff_new %>%
  group_by(country, gear) %>%
  summarise(
    nom_eff = sum(nom_active, na.rm = TRUE),
    nv_act  = sum(nv_active, na.rm = TRUE),
    p_act   = sum(p_active,  na.rm = TRUE),
    mean_power = p_act / nv_act,
    mean_das   = nom_eff / p_act,
    .groups = "drop"
  ) %>%
  group_by(country) %>%
  mutate(share_nom_eff = nom_eff / sum(nom_eff),
         share_nv = nv_act/sum(nv_act)) %>%
  ungroup() %>%
  arrange(country, desc(nom_eff))

# AUS: ~35% of AUS effort comes from midwater trawls, ~14% from Longlines, ~18% from "other", ~8% from handlines and poles, ~11% from seiners, ~6% gillnets, ~4% from pots and traps, ~2% from bottom trawl, ~1% dredges...
# NZL: ~33% from midwater trawls, ~33% from gillnets, ~21% from Others, ~7.6% handlines and poles, ~4 from seiners, <1% from dredges

gear_shares <- gear_totals %>%
  select(country, gear, share_nom_eff) %>%
  tidyr::pivot_wider(names_from = country, values_from = share_nom_eff)

aus_total <- country_decomp %>% filter(country == "AUS") %>% pull(nom_eff_total)

counterfactual_aus_by_gear <- gear_shares %>%
  mutate(
    aus_expected_if_nz_shares = aus_total * NZL
  )
counterfactual_aus_by_gear
```



