---
title: "Industrial validation with sentinel-1"
output: html_document
date: "2025-10-10"
---

# Validation Analysis: 
 - Compare industrial vessel predictions with Sentinel-1 detections


```{r setup}
library(tidyverse)
library(caret)
library(ggplot2)
library(gridExtra)
library(here)

# Set plotting theme
theme_set(theme_minimal())
```

```{r fxns}

# Load data
load_data <- function() {
  cat("Loading data...\n")
  
  # Load vessel predictions
  predictions <- read_csv(here('data/validation/ind_fishing_s1_area_presence.csv'), show_col_types = FALSE)
  cat(sprintf("Loaded %d vessel prediction records\n", nrow(predictions)))
  
  # Load Sentinel-1 detections
  s1_detections <- read_csv(here('data/validation/s1_presence_1deg.csv'), show_col_types = FALSE)
  cat(sprintf("Loaded %d Sentinel-1 detection records\n", nrow(s1_detections)))
  
  return(list(predictions = predictions, s1_detections = s1_detections))
}

# Create study area grid
create_study_area_grid <- function() {
  cat("\nLoading study area grid...\n")
  
  # Load the predefined study area
  study_area <- read_csv(here('data/validation/paolo_study_area_1deg.csv'), show_col_types = FALSE)
  all_locs <- study_area %>% select(lon, lat)
  
  cat(sprintf("Study area contains %d unique 1-degree grid cells\n", nrow(all_locs)))
  cat(sprintf("Latitude range: (%d, %d)\n", min(all_locs$lat), max(all_locs$lat)))
  cat(sprintf("Longitude range: (%d, %d)\n", min(all_locs$lon), max(all_locs$lon)))
  
  return(all_locs)
}

# Merge datasets
merge_datasets <- function(predictions, s1_detections, study_grid) {
  cat("\nMerging datasets...\n")
  
  # Prepare prediction data
  pred_data <- predictions %>%
    select(lon, lat, presence) %>%
    rename(predicted = presence)
  
  # Prepare S1 detection data
  s1_data <- s1_detections %>%
    select(lon, lat, presence) %>%
    rename(observed = presence)
  
  # Merge with study grid
  merged <- study_grid %>%
    left_join(pred_data, by = c("lon", "lat")) %>%
    left_join(s1_data, by = c("lon", "lat")) %>%
    mutate(
      predicted = replace_na(predicted, 0),
      observed = replace_na(observed, 0)
    )
  
  cat(sprintf("Total grid cells: %d\n", nrow(merged)))
  cat(sprintf("Cells with predictions: %d\n", sum(merged$predicted)))
  cat(sprintf("Cells with S1 detections: %d\n", sum(merged$observed)))
  
  return(merged)
}

```

Calculate validation metrics

```{r}
calculate_metrics <- function(merged_data) {
  cat("\n============================================================\n")
  cat("VALIDATION METRICS\n")
  cat("============================================================\n\n")

  # Ensure factors are exactly c("0","1") with "1" as the positive class
  pred_f <- factor(merged_data$predicted, levels = c(0, 1), labels = c("0", "1"))
  obs_f  <- factor(merged_data$observed,  levels = c(0, 1), labels = c("0", "1"))

  # Confusion table with rows = Observed (reference), cols = Predicted (data)
  cm_tbl <- table(Observed = obs_f, Predicted = pred_f)

  # Extract counts deterministically
  # Rows: Observed 0,1 ; Cols: Predicted 0,1
  tn <- as.integer(cm_tbl["0", "0"])
  fp <- as.integer(cm_tbl["0", "1"])
  fn <- as.integer(cm_tbl["1", "0"])
  tp <- as.integer(cm_tbl["1", "1"])

  total <- tn + fp + fn + tp

  # Metrics computed directly from counts
  accuracy    <- (tp + tn) / total
  precision   <- if ((tp + fp) > 0) tp / (tp + fp) else NA_real_
  recall      <- if ((tp + fn) > 0) tp / (tp + fn) else NA_real_         # sensitivity
  specificity <- if ((tn + fp) > 0) tn / (tn + fp) else NA_real_
  f1 <- if (is.na(precision) || is.na(recall) || (precision + recall) == 0) {
    NA_real_
  } else {
    2 * precision * recall / (precision + recall)
  }

  # Print a neat table
  cat("Confusion Matrix (rows=Observed, cols=Predicted):\n")
  print(cm_tbl)
  cat("\nCounts:\n")
  cat(sprintf("  TP: %d  TN: %d  FP: %d  FN: %d\n", tp, tn, fp, fn))

  cat(sprintf("\nOverall Metrics:\n"))
  cat(sprintf("  Accuracy:    %.4f\n", accuracy))
  cat(sprintf("  Precision:   %.4f (TP / (TP + FP))\n", precision))
  cat(sprintf("  Recall:      %.4f (TP / (TP + FN))\n", recall))
  cat(sprintf("  F1 Score:    %.4f\n", f1))
  cat(sprintf("  Specificity: %.4f (TN / (TN + FP))\n", specificity))

  # Agreement analysis
  both_present <- tp
  both_absent  <- tn
  pred_only    <- fp
  obs_only     <- fn

  cat(sprintf("\nSpatial Agreement:\n"))
  cat(sprintf("  Both present (TP):     %5d (%.2f%%)\n", both_present, 100 * both_present / total))
  cat(sprintf("  Both absent (TN):      %5d (%.2f%%)\n", both_absent,  100 * both_absent  / total))
  cat(sprintf("  Predicted only (FP):   %5d (%.2f%%)\n", pred_only,    100 * pred_only    / total))
  cat(sprintf("  Observed only (FN):    %5d (%.2f%%)\n", obs_only,     100 * obs_only     / total))
  cat(sprintf("  Total agreement:       %5d (%.2f%%)\n", both_present + both_absent,
              100 * (both_present + both_absent) / total))

  # Return everything the rest of your pipeline expects
  return(list(
    confusion_matrix = cm_tbl,
    accuracy = as.numeric(accuracy),
    precision = as.numeric(precision),
    recall = as.numeric(recall),
    f1 = as.numeric(f1),
    specificity = as.numeric(specificity),
    tp = tp, tn = tn, fp = fp, fn = fn
  ))
}


```

Confusion matrix

```{r}
plot_confusion_matrix <- function(cm, save_path = ('figs/validation/confusion_matrix.png')) {
  cm_df <- as.data.frame(cm)
  colnames(cm_df) <- c("Observed", "Predicted", "Count")
  
  cm_df$Observed <- factor(cm_df$Observed, levels = c(1, 0), labels = c("Present", "Absent"))
  cm_df$Predicted <- factor(cm_df$Predicted, levels = c(1, 0), labels = c("Present", "Absent"))
  
  p <- ggplot(cm_df, aes(x = Predicted, y = Observed, fill = Count)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Count), size = 10, color = "black") +
    scale_fill_gradient(low = "white", high = "steelblue") +
    labs(
      title = "Confusion Matrix: Model Predictions vs Sentinel-1 Detections",
      x = "Predicted",
      y = "Observed (Sentinel-1)"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 13, face = "bold")
    )
  
  ggsave(save_path, p, width = 8, height = 6, dpi = 300)
  cat(sprintf("\nSaved confusion matrix plot to %s\n", save_path))
}


```

Spatial comparison

```{r}
plot_spatial_comparison <- function(merged_data, save_path = 'figs/validation/spatial_comparison.png') {
  # Create agreement categories
  merged_data <- merged_data %>%
    mutate(
      category = case_when(
        predicted == 1 & observed == 1 ~ "Both (TP)",
        predicted == 1 & observed == 0 ~ "Predicted only (FP)",
        predicted == 0 & observed == 1 ~ "Observed only (FN)",
        TRUE ~ "No fishing (TN)"
      )
    )
  
  # Plot 1: Predictions
  p1 <- merged_data %>%
    filter(predicted == 1) %>%
    ggplot(aes(x = lon, y = lat)) +
    geom_point(color = "#2E2585", size = 0.5, alpha = 0.6) +
    labs(
      title = "A. Model Predictions",
      subtitle = sprintf("n = %d", sum(merged_data$predicted == 1)),
      x = "Longitude",
      y = "Latitude"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 20), 
              plot.subtitle = element_text(
      size = 18),
          # Axis titles
    axis.title.x = element_text(size = 18),  # x-axis label
    axis.title.y = element_text(size = 18),  # y-axis label

    # Axis tick labels
    axis.text.x  = element_text(size = 16),  # x tick text
    axis.text.y  = element_text(size = 16))   # y tick text)
  
  # Plot 2: Observations
  p2 <- merged_data %>%
    filter(observed == 1) %>%
    ggplot(aes(x = lon, y = lat)) +
    geom_point(color = "#C26D77", size = 0.5, alpha = 0.6) +
    labs(
      title = "B. Sentinel-1 Detections",
      subtitle = sprintf("n = %d", sum(merged_data$observed == 1)),
      x = "Longitude",
      y = "Latitude"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 20), 
              plot.subtitle = element_text(
      size = 18),
          # Axis titles
    axis.title.x = element_text(size = 18),  # x-axis label
    axis.title.y = element_text(size = 18),  # y-axis label

    # Axis tick labels
    axis.text.x  = element_text(size = 16),  # x tick text
    axis.text.y  = element_text(size = 16))   # y tick text)
  
  cat_labs <- merged_data %>%
  count(category) %>%
  mutate(label = paste0(category, " \n(n=", n, ")"))

  
  lab_vec <- setNames(cat_labs$label, cat_labs$category)

p3 <- merged_data %>%
  ggplot(aes(x = lon, y = lat, color = category)) +
  geom_point(size = 1, alpha = 0.7) +
  scale_color_manual(
    values = c(
      "Both (TP)"          = "#337538",
      "Predicted only (FP)" = "#DCCD7D",
      "Observed only (FN)"  = "#C26D77",
      "No fishing (TN)"     = "#DDDDDD"
    ),
    breaks = names(lab_vec),
    labels = lab_vec
  ) +
  labs(
    title = "C. Spatial Agreement: Predictions vs Observations",
    x = "Longitude",
    y = "Latitude",
    color = "Category"
  ) +
  theme_minimal() +
  theme(
    plot.title   = element_text(face = "bold", size = 20),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x  = element_text(size = 16),
    axis.text.y  = element_text(size = 16),

    legend.position    = "bottom",
    legend.direction   = "horizontal",
    legend.title       = element_text(size = 16, face = "bold"),
    legend.text        = element_text(size = 16),
    legend.key.width   = unit(1.6, "cm"),
    legend.key.height  = unit(0.7, "cm"),
    legend.spacing.x   = unit(0.6, "cm")
  ) +
  guides(
    color = guide_legend(
      nrow = 2,           # split legend into 2 rows -> less crowding
      byrow = TRUE,
      override.aes = list(
        size  = 4.5,      # bigger legend symbols
        alpha = 1
      )
    )
  )

  # Combine plots
  combined <- grid.arrange(p1, p2, p3, 
                          layout_matrix = rbind(c(1, 2), c(3, 3)),
                          heights = c(1, 1.2))
  
  ggsave(save_path, combined, width = 16, height = 12, dpi = 300)
  cat(sprintf("Saved spatial comparison plot to %s\n", save_path))
}

```


Metrics 

```{r}
plot_metrics_summary <- function(metrics, save_path = 'figs/validation/metrics_summary.png') {
  # Metrics bar plot
  metrics_df <- data.frame(
    Metric = c("Accuracy", "Precision", "Recall", "F1 Score", "Specificity"),
    Value = c(metrics$accuracy, metrics$precision, metrics$recall, 
              metrics$f1, metrics$specificity)
  )
  
  p1 <- ggplot(metrics_df, aes(x = reorder(Metric, Value), y = Value, fill = Metric)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = sprintf("%.3f", Value)), hjust = -0.1, size = 3.5) +
    coord_flip() +
    ylim(0, 1.1) +
    scale_fill_brewer(palette = "Set2") +
    labs(
      title = "Validation Metrics",
      x = "",
      y = "Score"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.text = element_text(size = 11)
    )
  
  # Confusion matrix breakdown
  cm_df <- data.frame(
    Category = c("True\nNegatives", "False\nPositives", "False\nNegatives", "True\nPositives"),
    Count = c(metrics$tn, metrics$fp, metrics$fn, metrics$tp),
    Color = c("#2ca02c", "#ff7f0e", "#d62728", "#1f77b4")
  )
  
  p2 <- ggplot(cm_df, aes(x = Category, y = Count, fill = Color)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = format(Count, big.mark = ",")), vjust = -0.5, size = 3.5) +
    scale_fill_identity() +
    labs(
      title = "Confusion Matrix Breakdown",
      x = "",
      y = "Count"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.text = element_text(size = 11)
    )
  
  # Combine plots
  combined <- grid.arrange(p1, p2, ncol = 2)
  
  ggsave(save_path, combined, width = 14, height = 5, dpi = 300)
  cat(sprintf("Saved metrics summary plot to %s\n", save_path))
}
```

Results

```{r}
export_results <- function(merged_data, metrics) {
  # Create results directory
  if (!dir.exists('results')) {
    dir.create('results')
  }
  
  # Export merged data with agreement categories
  merged_export <- merged_data %>%
    mutate(
      agreement = case_when(
        predicted == 1 & observed == 1 ~ "TP",
        predicted == 1 & observed == 0 ~ "FP",
        predicted == 0 & observed == 1 ~ "FN",
        predicted == 0 & observed == 0 ~ "TN"
      )
    )
  
  write_csv(merged_export, 'results/validation_results.csv')
  cat("\nExported detailed results to results/validation_results.csv\n")
  
  # Export metrics summary
  metrics_summary <- data.frame(
    Metric = c("Accuracy", "Precision", "Recall", "F1_Score", "Specificity",
               "True_Positives", "True_Negatives", "False_Positives", "False_Negatives"),
    Value = c(metrics$accuracy, metrics$precision, metrics$recall, metrics$f1,
              metrics$specificity, metrics$tp, metrics$tn, metrics$fp, metrics$fn)
  )
  
  write_csv(metrics_summary, 'results/metrics_summary.csv')
  cat("Exported metrics summary to results/metrics_summary.csv\n")
}
```

Run functions

```{r}
main <- function() {
  # Create output directories
  if (!dir.exists('figs')) {
    dir.create('figs')
  }
  if (!dir.exists('results')) {
    dir.create('results')
  }
  
  # Load data
  data <- load_data()
  
  # Create study area grid
  study_grid <- create_study_area_grid()
  
  # Merge datasets
  merged_data <- merge_datasets(data$predictions, data$s1_detections, study_grid)
  
  # Calculate metrics
  metrics <- calculate_metrics(merged_data)
  
  # Create visualizations
  cat("\nGenerating visualizations...\n")
  plot_confusion_matrix(metrics$confusion_matrix)
  plot_spatial_comparison(merged_data)
  plot_metrics_summary(metrics)
  
  # Export results
  export_results(merged_data, metrics)
  
  cat("\n============================================================\n")
  cat("VALIDATION ANALYSIS COMPLETE\n")
  cat("============================================================\n\n")
  cat("Output files created:\n")
  cat("  - figs/confusion_matrix.png\n")
  cat("  - figs/spatial_comparison.png\n")
  cat("  - figs/metrics_summary.png\n")
  cat("  - results/validation_results.csv\n")
  cat("  - results/metrics_summary.csv\n")
}

# Run main function
main()

```

## Figure 4

```{r}
  # Load data
  data <- load_data()
  
  # Create study area grid
  study_grid <- create_study_area_grid()
  
  # Merge datasets
  merged_data <- merge_datasets(data$predictions, data$s1_detections, study_grid)
  
    merged_data <- merged_data %>%
    mutate(
      category = case_when(
        predicted == 1 & observed == 1 ~ "Both (TP)",
        predicted == 1 & observed == 0 ~ "Predicted only (FP)",
        predicted == 0 & observed == 1 ~ "Observed only (FN)",
        TRUE ~ "No fishing (TN)"
      )
    )
  
  # Plot 1: Predictions
  p1 <- merged_data %>%
    filter(predicted == 1) %>%
    ggplot(aes(x = lon, y = lat)) +
    geom_point(color = "#2E2585", size = 0.5, alpha = 0.6) +
    labs(
      title = "A. Model Predictions",
      subtitle = sprintf("n = %d", sum(merged_data$predicted == 1)),
      x = "Longitude",
      y = "Latitude"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 20), 
              plot.subtitle = element_text(
      size = 18),
          # Axis titles
    axis.title.x = element_text(size = 18),  # x-axis label
    axis.title.y = element_text(size = 18),  # y-axis label

    # Axis tick labels
    axis.text.x  = element_text(size = 16),  # x tick text
    axis.text.y  = element_text(size = 16))   # y tick text)
  
  # Plot 2: Observations
  p2 <- merged_data %>%
    filter(observed == 1) %>%
    ggplot(aes(x = lon, y = lat)) +
    geom_point(color = "#C26D77", size = 0.5, alpha = 0.6) +
    labs(
      title = "B. Sentinel-1 Detections",
      subtitle = sprintf("n = %d", sum(merged_data$observed == 1)),
      x = "Longitude",
      y = "Latitude"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 20), 
              plot.subtitle = element_text(
      size = 18),
          # Axis titles
    axis.title.x = element_text(size = 18),  # x-axis label
    axis.title.y = element_text(size = 18),  # y-axis label

    # Axis tick labels
    axis.text.x  = element_text(size = 16),  # x tick text
    axis.text.y  = element_text(size = 16))   # y tick text)
  
  cat_labs <- merged_data %>%
  count(category) %>%
  mutate(label = paste0(category, " \n(n=", n, ")"))

  
  lab_vec <- setNames(cat_labs$label, cat_labs$category)

p3 <- merged_data %>%
  ggplot(aes(x = lon, y = lat, color = category)) +
  geom_point(size = 1, alpha = 0.7) +
  scale_color_manual(
    values = c(
      "Both (TP)"          = "#337538",
      "Predicted only (FP)" = "#DCCD7D",
      "Observed only (FN)"  = "#C26D77",
      "No fishing (TN)"     = "#DDDDDD"
    ),
    breaks = names(lab_vec),
    labels = lab_vec
  ) +
  labs(
    title = "C. Spatial Agreement: Predicted vs Observed",
    x = "Longitude",
    y = "Latitude",
    color = "Category"
  ) +
  theme_minimal() +
  theme(
    plot.title   = element_text(face = "bold", size = 20),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x  = element_text(size = 16),
    axis.text.y  = element_text(size = 16),

    legend.position    = "bottom",
    legend.direction   = "horizontal",
    legend.title       = element_text(size = 16, face = "bold"),
    legend.text        = element_text(size = 16),
    legend.key.width   = unit(1.6, "cm"),
    legend.key.height  = unit(0.7, "cm"),
    legend.spacing.x   = unit(0.6, "cm")
  ) +
  guides(
    color = guide_legend(
      nrow = 2,           # split legend into 2 rows -> less crowding
      byrow = TRUE,
      override.aes = list(
        size  = 4.5,      # bigger legend symbols
        alpha = 1
      )
    )
  )


  metrics <- calculate_metrics(merged_data)

  # Metrics bar plot
  metrics_df <- data.frame(
    Metric = c("Accuracy", "Precision", "Recall", "F1 Score", "Specificity"),
    Value = c(metrics$accuracy, metrics$precision, metrics$recall, 
              metrics$f1, metrics$specificity)
  )
 
cb_pal <- c(
  "Accuracy"     = "#0072B2",  # blue
  "Precision"    = "#E69F00",  # orange
  "Recall"       = "#009E73",  # green
  "F1 Score"     = "#CC79A7",  # purple/pink
  "Specificity"  = "#D55E00"   # vermillion
)

p4 <- ggplot(metrics_df, aes(x = reorder(Metric, Value), y = Value, fill = Metric)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.3f", Value)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  ylim(0, 1.1) +
  scale_fill_manual(values = cb_pal, drop = FALSE) +
  labs(
    title = "D. Validation Metrics",
    x = "",
    y = "Score"
  ) +
  theme_minimal() +
  theme(
    plot.title   = element_text(face = "bold", size = 20),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x  = element_text(size = 16),
    axis.text.y  = element_text(size = 16)
  )

  combined <- grid.arrange(
  p1, p2, p3, p4,
  ncol = 2, nrow = 2
)
  
  ggsave(here("prep/06_manuscript/figs/fig4_validation_plot_combined.png"), combined, width = 16, height = 12, dpi = 300)

```



Calculate balanced accuracy

```{r}
metrics_summary <- read.csv("results/metrics_summary.csv")

validation_results <- read.csv("results/validation_results.csv")

# balanced accuracy = (Sensitivity + Specificity)/2 = (TP/(TP+FN) + TN/(TN+FP))/2

TP <- metrics_summary %>% filter(Metric == "True_Positives") %>% pull(Value)

TN <- metrics_summary %>% filter(Metric == "True_Negatives") %>% pull(Value)

FN <- metrics_summary %>% filter(Metric == "False_Negatives") %>% pull(Value)

FP <- metrics_summary %>% filter(Metric == "False_Positives") %>% pull(Value)

balanced_accuracy = ((TP/(TP+FN)) + (TN/(TN+FP)))/2 # 0.65

mcc_num <- TP*TN - FP*FN
mcc_den <- sqrt((TP+FP)*(TP+FN)*(TN+FP)*(TN+FN))
mcc <- mcc_num / mcc_den
mcc # 0.344

study_area <- read_csv(here('data/validation/paolo_study_area_1deg.csv'), show_col_types = FALSE)

p2 <- study_area %>%
    ggplot(aes(x = lon, y = lat)) +
    geom_point(color = "red", size = 0.5, alpha = 0.6) +
    labs(
      title = "Study area",
      x = "Longitude",
      y = "Latitude"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold"))

```




Now run the same analysis, but use the rousseau data we prepped

```{r}

# Load data
load_data <- function() {
  cat("Loading data...\n")
  
  # Load vessel predictions
  predictions <- read_csv(here('data/validation/rousseau_fishing_s1_area_presence.csv'), show_col_types = FALSE)
  cat(sprintf("Loaded %d vessel prediction records\n", nrow(predictions)))
  
  # Load Sentinel-1 detections
  s1_detections <- read_csv(here('data/validation/s1_presence_1deg.csv'), show_col_types = FALSE)
  cat(sprintf("Loaded %d Sentinel-1 detection records\n", nrow(s1_detections)))
  
  return(list(predictions = predictions, s1_detections = s1_detections))
}

# Create study area grid
create_study_area_grid <- function() {
  cat("\nLoading study area grid...\n")
  
  # Load the predefined study area
  study_area <- read_csv(here('data/validation/paolo_study_area_1deg.csv'), show_col_types = FALSE)
  all_locs <- study_area %>% select(lon, lat)
  
  cat(sprintf("Study area contains %d unique 1-degree grid cells\n", nrow(all_locs)))
  cat(sprintf("Latitude range: (%d, %d)\n", min(all_locs$lat), max(all_locs$lat)))
  cat(sprintf("Longitude range: (%d, %d)\n", min(all_locs$lon), max(all_locs$lon)))
  
  return(all_locs)
}

# Merge datasets
merge_datasets <- function(predictions, s1_detections, study_grid) {
  cat("\nMerging datasets...\n")
  
  # Prepare prediction data
  pred_data <- predictions %>%
    select(lon, lat, presence) %>%
    rename(predicted = presence)
  
  # Prepare S1 detection data
  s1_data <- s1_detections %>%
    select(lon, lat, presence) %>%
    rename(observed = presence)
  
  # Merge with study grid
  merged <- study_grid %>%
    left_join(pred_data, by = c("lon", "lat")) %>%
    left_join(s1_data, by = c("lon", "lat")) %>%
    mutate(
      predicted = replace_na(predicted, 0),
      observed = replace_na(observed, 0)
    )
  
  cat(sprintf("Total grid cells: %d\n", nrow(merged)))
  cat(sprintf("Cells with predictions: %d\n", sum(merged$predicted)))
  cat(sprintf("Cells with S1 detections: %d\n", sum(merged$observed)))
  
  return(merged)
}

```

Calculate validation metrics

```{r}
calculate_metrics <- function(merged_data) {
  cat("\n============================================================\n")
  cat("VALIDATION METRICS\n")
  cat("============================================================\n\n")

  # Ensure factors are exactly c("0","1") with "1" as the positive class
  pred_f <- factor(merged_data$predicted, levels = c(0, 1), labels = c("0", "1"))
  obs_f  <- factor(merged_data$observed,  levels = c(0, 1), labels = c("0", "1"))

  # Confusion table with rows = Observed (reference), cols = Predicted (data)
  cm_tbl <- table(Observed = obs_f, Predicted = pred_f)

  # Extract counts deterministically
  # Rows: Observed 0,1 ; Cols: Predicted 0,1
  tn <- as.integer(cm_tbl["0", "0"])
  fp <- as.integer(cm_tbl["0", "1"])
  fn <- as.integer(cm_tbl["1", "0"])
  tp <- as.integer(cm_tbl["1", "1"])

  total <- tn + fp + fn + tp

  # Metrics computed directly from counts
  accuracy    <- (tp + tn) / total
  precision   <- if ((tp + fp) > 0) tp / (tp + fp) else NA_real_
  recall      <- if ((tp + fn) > 0) tp / (tp + fn) else NA_real_         # sensitivity
  specificity <- if ((tn + fp) > 0) tn / (tn + fp) else NA_real_
  f1 <- if (is.na(precision) || is.na(recall) || (precision + recall) == 0) {
    NA_real_
  } else {
    2 * precision * recall / (precision + recall)
  }

  # Print a neat table
  cat("Confusion Matrix (rows=Observed, cols=Modeled):\n")
  print(cm_tbl)
  cat("\nCounts:\n")
  cat(sprintf("  TP: %d  TN: %d  FP: %d  FN: %d\n", tp, tn, fp, fn))

  cat(sprintf("\nOverall Metrics:\n"))
  cat(sprintf("  Accuracy:    %.4f\n", accuracy))
  cat(sprintf("  Precision:   %.4f (TP / (TP + FP))\n", precision))
  cat(sprintf("  Recall:      %.4f (TP / (TP + FN))\n", recall))
  cat(sprintf("  F1 Score:    %.4f\n", f1))
  cat(sprintf("  Specificity: %.4f (TN / (TN + FP))\n", specificity))

  # Agreement analysis
  both_present <- tp
  both_absent  <- tn
  pred_only    <- fp
  obs_only     <- fn

  cat(sprintf("\nSpatial Agreement:\n"))
  cat(sprintf("  Both present (TP):     %5d (%.2f%%)\n", both_present, 100 * both_present / total))
  cat(sprintf("  Both absent (TN):      %5d (%.2f%%)\n", both_absent,  100 * both_absent  / total))
  cat(sprintf("  Modeled only (FP):   %5d (%.2f%%)\n", pred_only,    100 * pred_only    / total))
  cat(sprintf("  Observed only (FN):    %5d (%.2f%%)\n", obs_only,     100 * obs_only     / total))
  cat(sprintf("  Total agreement:       %5d (%.2f%%)\n", both_present + both_absent,
              100 * (both_present + both_absent) / total))

  # Return everything the rest of your pipeline expects
  return(list(
    confusion_matrix = cm_tbl,
    accuracy = as.numeric(accuracy),
    precision = as.numeric(precision),
    recall = as.numeric(recall),
    f1 = as.numeric(f1),
    specificity = as.numeric(specificity),
    tp = tp, tn = tn, fp = fp, fn = fn
  ))
}

```



Confusion matrix

```{r}
plot_confusion_matrix <- function(cm, save_path = ('figs/validation/confusion_matrix_rousseau.png')) {
  cm_df <- as.data.frame(cm)
  colnames(cm_df) <- c("Observed", "Modeled", "Count")
  
  cm_df$Observed <- factor(cm_df$Observed, levels = c(1, 0), labels = c("Present", "Absent"))
  cm_df$Predicted <- factor(cm_df$Modeled, levels = c(1, 0), labels = c("Present", "Absent"))
  
  p <- ggplot(cm_df, aes(x = Modeled, y = Observed, fill = Count)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Count), size = 10, color = "black") +
    scale_fill_gradient(low = "white", high = "steelblue") +
    labs(
      title = "Confusion Matrix: Model Predictions vs Sentinel-1 Detections",
      x = "Modeled",
      y = "Observed (Sentinel-1)"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 13, face = "bold")
    )
  
  ggsave(save_path, p, width = 8, height = 6, dpi = 300)
  cat(sprintf("\nSaved confusion matrix plot to %s\n", save_path))
}


```

Spatial comparison

```{r}
plot_spatial_comparison <- function(merged_data, save_path = 'figs/validation/spatial_comparison_rousseau.png') {
  # Create agreement categories
  merged_data <- merged_data %>%
    mutate(
      category = case_when(
        predicted == 1 & observed == 1 ~ "Both (TP)",
        predicted == 1 & observed == 0 ~ "Modeled only (FP)",
        predicted == 0 & observed == 1 ~ "Observed only (FN)",
        TRUE ~ "No fishing (TN)"
      )
    )
  
  # Plot 1: Predictions
  p1 <- merged_data %>%
    filter(predicted == 1) %>%
    ggplot(aes(x = lon, y = lat)) +
    geom_point(color = "#2E2585", size = 0.5, alpha = 0.6) +
    labs(
      title = "A. Rousseau Model",
      subtitle = sprintf("n = %d", sum(merged_data$predicted == 1)),
      x = "Longitude",
      y = "Latitude"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 20), 
              plot.subtitle = element_text(
      size = 18),
          # Axis titles
    axis.title.x = element_text(size = 18),  # x-axis label
    axis.title.y = element_text(size = 18),  # y-axis label

    # Axis tick labels
    axis.text.x  = element_text(size = 16),  # x tick text
    axis.text.y  = element_text(size = 16))   # y tick text)
  
  # Plot 2: Observations
  p2 <- merged_data %>%
    filter(observed == 1) %>%
    ggplot(aes(x = lon, y = lat)) +
    geom_point(color = "#C26D77", size = 0.5, alpha = 0.6) +
    labs(
      title = "B. Sentinel-1 Detections",
      subtitle = sprintf("n = %d", sum(merged_data$observed == 1)),
      x = "Longitude",
      y = "Latitude"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 20), 
              plot.subtitle = element_text(
      size = 18),
          # Axis titles
    axis.title.x = element_text(size = 18),  # x-axis label
    axis.title.y = element_text(size = 18),  # y-axis label

    # Axis tick labels
    axis.text.x  = element_text(size = 16),  # x tick text
    axis.text.y  = element_text(size = 16))   # y tick text)
  
  cat_labs <- merged_data %>%
  count(category) %>%
  mutate(label = paste0(category, " \n(n=", n, ")"))

  
  lab_vec <- setNames(cat_labs$label, cat_labs$category)

p3 <- merged_data %>%
  ggplot(aes(x = lon, y = lat, color = category)) +
  geom_point(size = 1, alpha = 0.7) +
  scale_color_manual(
    values = c(
      "Both (TP)"          = "#337538",
      "Modeled only (FP)" = "#DCCD7D",
      "Observed only (FN)"  = "#C26D77",
      "No fishing (TN)"     = "#DDDDDD"
    ),
    breaks = names(lab_vec),
    labels = lab_vec
  ) +
  labs(
    title = "C. Spatial Agreement: Rousseau model vs Observations",
    x = "Longitude",
    y = "Latitude",
    color = "Category"
  ) +
  theme_minimal() +
  theme(
    plot.title   = element_text(face = "bold", size = 20),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x  = element_text(size = 16),
    axis.text.y  = element_text(size = 16),

    legend.position    = "bottom",
    legend.direction   = "horizontal",
    legend.title       = element_text(size = 16, face = "bold"),
    legend.text        = element_text(size = 16),
    legend.key.width   = unit(1.6, "cm"),
    legend.key.height  = unit(0.7, "cm"),
    legend.spacing.x   = unit(0.6, "cm")
  ) +
  guides(
    color = guide_legend(
      nrow = 2,           # split legend into 2 rows -> less crowding
      byrow = TRUE,
      override.aes = list(
        size  = 4.5,      # bigger legend symbols
        alpha = 1
      )
    )
  )

  # Combine plots
  combined <- grid.arrange(p1, p2, p3, 
                          layout_matrix = rbind(c(1, 2), c(3, 3)),
                          heights = c(1, 1.2))
  
  ggsave(save_path, combined, width = 16, height = 12, dpi = 300)
  cat(sprintf("Saved spatial comparison plot to %s\n", save_path))
}

```

Metrics 

```{r}
plot_metrics_summary <- function(metrics, save_path = 'figs/validation/metrics_summary_rousseau.png') {
  # Metrics bar plot
  metrics_df <- data.frame(
    Metric = c("Accuracy", "Precision", "Recall", "F1 Score", "Specificity"),
    Value = c(metrics$accuracy, metrics$precision, metrics$recall, 
              metrics$f1, metrics$specificity)
  )
  
  p1 <- ggplot(metrics_df, aes(x = reorder(Metric, Value), y = Value, fill = Metric)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = sprintf("%.3f", Value)), hjust = -0.1, size = 3.5) +
    coord_flip() +
    ylim(0, 1.1) +
    scale_fill_brewer(palette = "Set2") +
    labs(
      title = "Validation Metrics",
      x = "",
      y = "Score"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.text = element_text(size = 11)
    )
  
  # Confusion matrix breakdown
  cm_df <- data.frame(
    Category = c("True\nNegatives", "False\nPositives", "False\nNegatives", "True\nPositives"),
    Count = c(metrics$tn, metrics$fp, metrics$fn, metrics$tp),
    Color = c("#2ca02c", "#ff7f0e", "#d62728", "#1f77b4")
  )
  
  p2 <- ggplot(cm_df, aes(x = Category, y = Count, fill = Color)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = format(Count, big.mark = ",")), vjust = -0.5, size = 3.5) +
    scale_fill_identity() +
    labs(
      title = "Confusion Matrix Breakdown",
      x = "",
      y = "Count"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.text = element_text(size = 11)
    )
  
  # Combine plots
  combined <- grid.arrange(p1, p2, ncol = 2)
  
  ggsave(save_path, combined, width = 14, height = 5, dpi = 300)
  cat(sprintf("Saved metrics summary plot to %s\n", save_path))
}
```

Results

```{r}
export_results <- function(merged_data, metrics) {
  # Create results directory
  if (!dir.exists('results')) {
    dir.create('results')
  }
  
  # Export merged data with agreement categories
  merged_export <- merged_data %>%
    mutate(
      agreement = case_when(
        predicted == 1 & observed == 1 ~ "TP",
        predicted == 1 & observed == 0 ~ "FP",
        predicted == 0 & observed == 1 ~ "FN",
        predicted == 0 & observed == 0 ~ "TN"
      )
    )
  
  write_csv(merged_export, 'results/validation_results_rousseau.csv')
  cat("\nExported detailed results to results/validation_results_rousseau.csv\n")
  
  # Export metrics summary
  metrics_summary <- data.frame(
    Metric = c("Accuracy", "Precision", "Recall", "F1_Score", "Specificity",
               "True_Positives", "True_Negatives", "False_Positives", "False_Negatives"),
    Value = c(metrics$accuracy, metrics$precision, metrics$recall, metrics$f1,
              metrics$specificity, metrics$tp, metrics$tn, metrics$fp, metrics$fn)
  )
  
  write_csv(metrics_summary, 'results/metrics_summary_rousseau.csv')
  cat("Exported metrics summary to results/metrics_summary_rousseau.csv\n")
}
```

Run functions

```{r}
main <- function() {
  # Create output directories
  if (!dir.exists('figs')) {
    dir.create('figs')
  }
  if (!dir.exists('results')) {
    dir.create('results')
  }
  
  # Load data
  data <- load_data()
  
  # Create study area grid
  study_grid <- create_study_area_grid()
  
  # Merge datasets
  merged_data <- merge_datasets(data$predictions, data$s1_detections, study_grid)
  
  # Calculate metrics
  metrics <- calculate_metrics(merged_data)
  
  # Create visualizations
  cat("\nGenerating visualizations...\n")
  plot_confusion_matrix(metrics$confusion_matrix)
  plot_spatial_comparison(merged_data)
  plot_metrics_summary(metrics)
  
  # Export results
  export_results(merged_data, metrics)
  
  cat("\n============================================================\n")
  cat("VALIDATION ANALYSIS COMPLETE\n")
  cat("============================================================\n\n")
  cat("Output files created:\n")
  cat("  - figs/confusion_matrix_rousseau.png\n")
  cat("  - figs/spatial_comparison_rousseau.png\n")
  cat("  - figs/metrics_summary_rousseau.png\n")
  cat("  - results/validation_results_rousseau.csv\n")
  cat("  - results/metrics_summary_rousseau.csv\n")
}

# Run main function
main()

```

## SI Figure 6

```{r}
  # Load data
  data <- load_data()
  
  # Create study area grid
  study_grid <- create_study_area_grid()
  
  # Merge datasets
  merged_data <- merge_datasets(data$predictions, data$s1_detections, study_grid)
  
    merged_data <- merged_data %>%
    mutate(
      category = case_when(
        predicted == 1 & observed == 1 ~ "Both (TP)",
        predicted == 1 & observed == 0 ~ "Modeled only (FP)",
        predicted == 0 & observed == 1 ~ "Observed only (FN)",
        TRUE ~ "No fishing (TN)"
      )
    )
  
  # Plot 1: Predictions
  p1 <- merged_data %>%
    filter(predicted == 1) %>%
    ggplot(aes(x = lon, y = lat)) +
    geom_point(color = "#2E2585", size = 0.5, alpha = 0.6) +
    labs(
      title = "A. Rousseau Model",
      subtitle = sprintf("n = %d", sum(merged_data$predicted == 1)),
      x = "Longitude",
      y = "Latitude"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 20), 
              plot.subtitle = element_text(
      size = 18),
          # Axis titles
    axis.title.x = element_text(size = 18),  # x-axis label
    axis.title.y = element_text(size = 18),  # y-axis label

    # Axis tick labels
    axis.text.x  = element_text(size = 16),  # x tick text
    axis.text.y  = element_text(size = 16))   # y tick text)
  
  # Plot 2: Observations
  p2 <- merged_data %>%
    filter(observed == 1) %>%
    ggplot(aes(x = lon, y = lat)) +
    geom_point(color = "#C26D77", size = 0.5, alpha = 0.6) +
    labs(
      title = "B. Sentinel-1 Detections",
      subtitle = sprintf("n = %d", sum(merged_data$observed == 1)),
      x = "Longitude",
      y = "Latitude"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 20), 
              plot.subtitle = element_text(
      size = 18),
          # Axis titles
    axis.title.x = element_text(size = 18),  # x-axis label
    axis.title.y = element_text(size = 18),  # y-axis label

    # Axis tick labels
    axis.text.x  = element_text(size = 16),  # x tick text
    axis.text.y  = element_text(size = 16))   # y tick text)
  
  cat_labs <- merged_data %>%
  count(category) %>%
  mutate(label = paste0(category, " \n(n=", n, ")"))

  
  lab_vec <- setNames(cat_labs$label, cat_labs$category)

p3 <- merged_data %>%
  ggplot(aes(x = lon, y = lat, color = category)) +
  geom_point(size = 1, alpha = 0.7) +
  scale_color_manual(
    values = c(
      "Both (TP)"          = "#337538",
      "Modeled only (FP)" = "#DCCD7D",
      "Observed only (FN)"  = "#C26D77",
      "No fishing (TN)"     = "#DDDDDD"
    ),
    breaks = names(lab_vec),
    labels = lab_vec
  ) +
  labs(
    title = "C. Spatial Agreement: Modeled vs Observed",
    x = "Longitude",
    y = "Latitude",
    color = "Category"
  ) +
  theme_minimal() +
  theme(
    plot.title   = element_text(face = "bold", size = 20),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x  = element_text(size = 16),
    axis.text.y  = element_text(size = 16),

    legend.position    = "bottom",
    legend.direction   = "horizontal",
    legend.title       = element_text(size = 16, face = "bold"),
    legend.text        = element_text(size = 16),
    legend.key.width   = unit(1.6, "cm"),
    legend.key.height  = unit(0.7, "cm"),
    legend.spacing.x   = unit(0.6, "cm")
  ) +
  guides(
    color = guide_legend(
      nrow = 2,           # split legend into 2 rows -> less crowding
      byrow = TRUE,
      override.aes = list(
        size  = 4.5,      # bigger legend symbols
        alpha = 1
      )
    )
  )


  metrics <- calculate_metrics(merged_data)

  # Metrics bar plot
  metrics_df <- data.frame(
    Metric = c("Accuracy", "Precision", "Recall", "F1 Score", "Specificity"),
    Value = c(metrics$accuracy, metrics$precision, metrics$recall, 
              metrics$f1, metrics$specificity)
  )
 
cb_pal <- c(
  "Accuracy"     = "#0072B2",  # blue
  "Precision"    = "#E69F00",  # orange
  "Recall"       = "#009E73",  # green
  "F1 Score"     = "#CC79A7",  # purple/pink
  "Specificity"  = "#D55E00"   # vermillion
)

p4 <- ggplot(metrics_df, aes(x = reorder(Metric, Value), y = Value, fill = Metric)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.3f", Value)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  ylim(0, 1.1) +
  scale_fill_manual(values = cb_pal, drop = FALSE) +
  labs(
    title = "D. Validation Metrics",
    x = "",
    y = "Score"
  ) +
  theme_minimal() +
  theme(
    plot.title   = element_text(face = "bold", size = 20),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x  = element_text(size = 16),
    axis.text.y  = element_text(size = 16)
  )

  combined <- grid.arrange(
  p1, p2, p3, p4,
  ncol = 2, nrow = 2
)
  
  ggsave(here("prep/06_manuscript/figs/SI_fig4_validation_plot_combined_rousseau.png"), combined, width = 16, height = 12, dpi = 300)

```