---
title: "Save data for Zenodo publishing"
output: html_document
date: "2024-12-11"
---

# Summary

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(terra)
library(foreach)
library(doParallel)
library(glue)
library(qs)
library(here)
library(countrycode)
library(strex)

source(here("R/dir.R"))
```


Save csv files to upload to Zenodo and the app 

 - We need to grab any gapfilled flags that are not in the flag level modelling from the regional level modelling 

```{r}

  modelled_flags_original <- str_before_first(str_after_last(list.files(file.path(rdsi_dir, "prep/random_forest/artisanal_skylight_sentinel/predictions/"), pattern = ".qs", full.names = TRUE), "_"), "\\.qs")

all_flags <- unique(modelled_flags_original)


cl <- makeCluster(24)  # could probably increase this? 
registerDoParallel(cl)


foreach(flag = all_flags, .packages = c("qs", "dplyr", "tidyverse", "glue", "countrycode")) %dopar% {
  # flag = "RAA"
  
  
  all_files_flag_original <- list.files(file.path(rdsi_dir, "prep/random_forest/artisanal_skylight_sentinel/predictions/"), pattern = flag, full.names = TRUE)
  # all_files_flag_regional_gf <- list.files(file.path(rdsi_dir, "prep/random_forest/predictions_regional/"), pattern = flag, full.names = TRUE)



  all_data_flag_original <- lapply(all_files_flag_original, qread) %>%
    bind_rows() %>%
    mutate(data_type = "flag model")

  
    if(nrow(all_data_flag_original) == 0){
    
    all_data_flag_original <- all_data_flag_regional_gf %>%
      mutate(nom_active_fishing_hours = 0, eff_active_fishing_hours = 0, nom_active_fishing_days = 0, eff_active_fishing_days = 0, eff_days_km2 = 0, nom_days_km2 = 0, eff_hours_km2 = 0, nom_hours_km2 = 0, data_type = NA)
    
  }
  
  all_data_fin <- all_data_flag_original %>%
    filter(nom_active_fishing_hours > 0) %>%
    mutate(flag_country_name = case_when(
      flag_country_iso3c == "RAA" ~ "Azores", 
      flag_country_iso3c == "RAM" ~ "Madeira", 
      TRUE ~ flag_country_name
    )) %>%
    mutate(flag_country_iso3c = flag)
  
  write.csv(all_data_fin, file.path(rdsi_dir, glue("prep/random_forest/zenodo_data/artisanal_skylight_sentinel/mapped_by_flag_country/model_preds_1950_2017_{flag}.csv")), row.names = FALSE)
}

stopCluster(cl)


## save as one large csv

  all_files_flag <- list.files(file.path(rdsi_dir, "prep/random_forest/zenodo_data/artisanal_skylight_sentinel/mapped_by_flag_country/"), full.names = TRUE)

    all_data <- lapply(all_files_flag, read.csv) %>%
      bind_rows()
    
    test <- all_data %>%
      filter(year == 2017) %>%
      group_by(lon, lat) %>% 
      summarise(effort = sum(nom_active_fishing_hours, na.rm = TRUE)) %>% 
      ungroup() %>%
      rast(., type = "xyz")
    
    plot(log(test+1))
    plot(test)
    
write.csv(all_data, file.path(rdsi_dir, "prep/random_forest/zenodo_data/artisanal_skylight_sentinel/mapped_artisanal_effort_predictions_1950_2017.csv"), row.names = FALSE)

all_data_grouped <- all_data %>% 
  group_by(year, flag_country_iso3c, flag_country_name, length_category, eez_sovereign_iso3c, eez_sovereign_name, fao_fishing_id, fao_major_fishing_area, sector) %>%
  summarise(eff_active_fishing_hours = sum(eff_active_fishing_hours, na.rm = TRUE), 
            eff_active_fishing_days = sum(eff_active_fishing_days, na.rm = TRUE), 
            nom_active_fishing_hours = sum(nom_active_fishing_hours, na.rm = TRUE), 
            nom_active_fishing_days = sum(nom_active_fishing_days, na.rm = TRUE),
            nv = sum(nv, na.rm = TRUE)) %>%
  ungroup()


write.csv(all_data_grouped, file.path(rdsi_dir, "prep/random_forest/zenodo_data/artisanal_skylight_sentinel/artisanal_effort_predictions_by_flag_eez_fao_gear_length_1950_2017.csv"), row.names = FALSE)



  all_data_grouped <- read.csv(file.path(rdsi_dir, "prep/random_forest/zenodo_data/artisanal_skylight_sentinel/artisanal_effort_predictions_by_flag_eez_fao_gear_length_1950_2017.csv"))
  
all_data_grouped_2 <- all_data_grouped %>% 
  group_by(year, flag_country_iso3c, flag_country_name, length_category) %>%
    summarise(modeled_eff_active_fishing_hours = sum(eff_active_fishing_hours, na.rm = TRUE), 
            modeled_eff_active_fishing_days = sum(eff_active_fishing_days, na.rm = TRUE), 
            modeled_nom_active_fishing_hours = sum(nom_active_fishing_hours, na.rm = TRUE), 
            modeled_nom_active_fishing_days = sum(nom_active_fishing_days, na.rm = TRUE),
            modeled_nv = sum(nv, na.rm = TRUE)) %>%
  ungroup() %>%
    mutate(flag_country_name = case_when(
          flag_country_iso3c == "RAM" ~ "Madeira",
          flag_country_iso3c == "RAA" ~ "Azores", 
          TRUE ~ flag_country_name
  ))

hist_fish_data <- qs::qread(here("data/model_features/rousseau_total_art_vessels.qs")) %>%
    dplyr::select(flag_country_iso3c = flag_fin, year, length_category, total_nv = nv, total_eff_days = eff_active_days, total_nom_days = nom_active_days) %>%
        filter(total_nv > 0) %>%
  mutate(flag_country_name = countrycode(flag_country_iso3c, origin = "iso3c", destination = "country.name")) %>%
  mutate(flag_country_name = case_when(
          flag_country_iso3c == "RAM" ~ "Madeira",
          flag_country_iso3c == "RAA" ~ "Azores", 
          TRUE ~ flag_country_name
  )) %>% 
  left_join(all_data_grouped_2) %>%
  mutate(proportion_nv_modeled = modeled_nv/total_nv,
         prop_nom_days_modeled = modeled_nom_active_fishing_days/total_nom_days)

missing_data <- hist_fish_data %>%
  filter(proportion_nv_modeled != 1) %>%
  filter(proportion_nv_modeled < 0.9999) # 0 awesome! 

write.csv(hist_fish_data, file.path(rdsi_dir, "prep/random_forest/zenodo_data/artisanal_skylight_sentinel/known_artisanal_effort_rousseau.csv"), row.names = FALSE)




test <- hist_fish_data %>% filter(flag_country_iso3c == "UKR", year == 1950)

sum(test$modeled_nom_active_fishing_hours) # 


```


```{r}
test <- hist_fish_data %>%
  filter(flag_country_iso3c %in% c("NOR", "NAM", "AUS", "NZL"))
```

Why does Norway stop in 1983? 
Australia stops in 2016? 
There is no New Zealand or Namibia at all? 

According to supplementary info from Rousseau et al. 2019 PNAS (https://www.pnas.org/doi/10.1073/pnas.1820344116#supplementary-materials):
 - Namibia has no artisanal fishing
 - Norway only has artisanal <12m until 2015? Feels like this is a mistake that wasn't corrected? 
 - New Zealand artisanal stopped in 2010? Also feels like a mistake. Seems like they should've carried over the definiton of artisanal to the latest year.. 
 - Australia artisanal is <10m and stopped in 2014? Definitely is not true. 

```{r}
library(janitor)
# read in published data
rousseau_eff <- read.csv("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/TotalEffortby_FishingCountry_LengthBoat_Gear_Sector.csv") %>% 
  clean_names() %>%
  filter(country %in% c("NOR", "NAM", "AUS", "NZL"),
         sector %in% c("UP", "APW"))

unique(rousseau_eff$sector) # UP only - weird
unique(rousseau_eff$country) # [1] "AUS" "NOR" - no NZL or NAM? 

test_years <- rousseau_eff %>%
  group_by(country, year) %>% 
  summarise(nv = sum(nv, na.rm = TRUE)) %>%
  ungroup()

# AUS to 2016, NOR to 1983 

## lets look at other published data from Rousseau and see if it is the same.. maybe it is a problem with the dataset we're using? 
options(timeout = 600)   # default is 60s; make it longer
mapped_2017_apw <- read.csv("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_2017_APW.csv")

mapped_2017_up <- read.csv("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_2017_UP.csv")
saup_codes <- read.csv("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/SAUPtoCountry.csv") %>%
  dplyr::select(-X)

mapped_2017_apw_up <- rbind(mapped_2017_apw, mapped_2017_up) %>%
  left_join(saup_codes)


# AUS = 36, 
# NZL = 554, 
# NOR = 578, 
# NAM = 516

test_missing <- mapped_2017_apw_up %>%
 # filter(SAUP %in% c(36, 554, 578, 516))
  filter(Country %in% c("NOR", "AUS", "NAM", "NZL")) # ok so these are all missing from 2017... just like the other dataset. I wonder if it just an artifact of what they are considering as artisanal? 

test_nam <- read.csv("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped_country/mapped_516.csv") # only industrial? Weird. According to google fisheries in Namibia are mostly industrial. 

test <- test_nam %>%
  group_by(Lon, Lat) %>%
  summarise(nom_eff = sum(NomActive, na.rm = TRUE)) %>%
  ungroup() %>%
  na.omit() %>%
 # filter(nom_eff > 1) %>%
  rast(., type = "xyz")

plot(test) # weird how spread out it is.. SAUP is much more concentrated in NAM EEZ, which makes sense. That is also how our industrial is. 

```

Lets compare our artisanal modelling to Rousseau et al. artisanal data 

```{r}
library(data.table)
saup_codes <- read.csv("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/SAUPtoCountry.csv")

# lets load in the mapped artisanal powered and unpowered and join
options(timeout = 600)  # extend timeout to 10 minutes

mapped_2017_apw <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_2017_APW.csv")

mapped_2017_up <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_2017_UP.csv")

all_rousseau_2017 <- rbind(mapped_2017_apw, mapped_2017_up)

  all_files_flag <- list.files(file.path(rdsi_dir, "prep/random_forest/zenodo_data/artisanal_skylight_sentinel/mapped_by_flag_country/"), full.names = TRUE)

    all_data <- lapply(all_files_flag, fread) %>%
      bind_rows()
    
    all_data_2017 <- all_data %>%
      filter(year == 2017)
    
test_new <- all_data_2017 %>%    
      group_by(lon, lat) %>%
      summarise(effort = sum(nom_active_fishing_hours, na.rm = TRUE)) %>%
      ungroup() %>%
      filter(effort > 0) %>%
      rast(., type = "xyz")
    
    
test_rousseau <- all_rousseau_2017 %>%
  group_by(Lon, Lat) %>%
  summarise(nom_eff = sum(NomActive, na.rm = TRUE)) %>%
  ungroup() %>%
  na.omit() %>%
  filter(nom_eff > 0) %>%
  rast(., type = "xyz")


par(mfrow = c(2, 1))  # 2 rows, 1 column
plot(test_new, main = "Predicted effort (new model)")
plot(test_rousseau, main = "Rousseau et al. 2019 effort")

# interesting, looks pretty similar, however, ours is definitely 1) more concentrated, and 2) closer to shore. I am a little confused because presumably Watson (and by extension Rousseau) have the same <=50km or >=200m depth as us, but it looks like their artisnal goes out to the edge of EEZs (very low effort all the way out on the edge though). Just looking at unpowered looks a lot more similar.

```

Lets make a plot showing effort in China in 1950 vs 1985 vs 2017 for industrial 

```{r}
library(scales)
library(here)

cell_area_rousseau <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/CellstoLME_EEZ.csv") %>%
  distinct(lon = Lon, lat = Lat, pixel_area_km2 = Area)

rousseau_ind_2017 <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_2017_I.csv")
rousseau_ind_1985 <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_1985_I.csv")
rousseau_ind_1950 <- fread("https://data.imas.utas.edu.au/attachments/1241a51d-c8c2-4432-aa68-3d2bae142794/effort_mapped/mapped_1950_I.csv")


rousseau_chn <- rousseau_ind_2017 %>%
  mutate(year = 2017) %>%
  filter(SAUP == 156) %>%
  rbind(rousseau_ind_1985 %>% filter(SAUP == 156) %>% mutate(year = 1985)) %>%
  rbind(rousseau_ind_1950 %>% filter(SAUP == 156) %>% mutate(year = 1950)) %>%
  group_by(year, lon = Lon, lat = Lat) %>%
  summarise(nom_eff = sum(NomActive, na.rm = TRUE)) %>%
  ungroup() %>%
  na.omit() %>%
  filter(nom_eff > 0) %>%
  transmute(year, lon, lat, nom_eff, source = "Rousseau et al 2024") %>%
  left_join(cell_area_rousseau) %>%
  mutate(eff_km2 = nom_eff/pixel_area_km2)
  
# SAUP = 156 for China
area_new <- fread(file.path(rdsi_dir, "prep/random_forest/zenodo_data/mapped_by_flag_country/model_preds_1950_2017_CHN.csv")) %>%
  distinct(lon, lat, pixel_area_km2)

new_china_ind <- fread(file.path(rdsi_dir, "prep/random_forest/zenodo_data/mapped_by_flag_country/model_preds_1950_2017_CHN.csv")) %>%
  filter(year %in% c(1950, 1985, 2017)) %>%
  group_by(year, lon, lat) %>%
  summarise(nom_eff = sum(nom_active_fishing_hours, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(nom_eff > 0) %>%
  transmute(year, lon, lat, nom_eff, source = "Model") %>%
  left_join(area_new) %>%
  mutate(eff_km2 = nom_eff/pixel_area_km2)

years <- c(1950, 1985, 2017)

pdat <- bind_rows(new_china_ind, rousseau_chn) %>%
  mutate(
    year   = factor(year, levels = years),
    source = factor(source, levels = c("Model", "Rousseau et al 2024"))
  )

# consistent extents across all facets
lims <- pdat %>%
  summarise(xmin = min(lon), xmax = max(lon),
            ymin = min(lat), ymax = max(lat))



# Your custom bins
breaks_num <- c(0.02, 0.2, 2, 20, 200, 2000)
labels     <- c("<0.02", "0.02-0.2", "0.2-2", "2-20", "20-200", "200-2000", ">2000")
colors     <- c("#FFFFE4", "#EFF3FE", "#CADBEE", "#A8C9E0", "#E8F4A2", "#F1B16D", "#C54B53")
names(colors) <- labels

# Combine old + new
pdat <- bind_rows(rousseau_chn, new_china_ind) %>%
  mutate(
    eff_bin = cut(
      eff_km2,
      breaks = c(-Inf, breaks_num, Inf),   # add open-ended bins
      labels = labels,
      right = FALSE,
      include.lowest = TRUE
    ),
    eff_bin = factor(eff_bin, levels = labels, ordered = TRUE)
  )

library(rnaturalearth)
library(rnaturalearthdata)
library(sf)

# 1. Get global land polygons (at medium resolution)
land <- ne_countries(scale = "medium", returnclass = "sf")

library(ggplot2)
library(dplyr)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)

land <- ne_countries(scale = "medium", returnclass = "sf")

# Plot with coord_sf()
p <- ggplot() +
  geom_raster(data = pdat, aes(lon, lat, fill = eff_bin)) +
  geom_sf(data = land, fill = "grey85", color = "grey60", linewidth = 0.1) +
  facet_grid(year ~ source) +
  coord_sf(
    xlim = range(pdat$lon, na.rm = TRUE),
    ylim = range(pdat$lat, na.rm = TRUE),
    expand = FALSE
  ) +
  scale_fill_manual(
    values = colors,
    drop = FALSE,
    name = "Fishing Effort (kW hours/km²)",
    guide = guide_legend(
      nrow = 1,
      byrow = TRUE,
      title.position = "top",
      title.hjust = 0.5
    )
  ) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )

# p
# 4. Save
ggsave(here("figs/fishing_effort_china_with_land.png"), plot = p, width = 10, height = 6, dpi = 600, bg = "white")



```




